---
title: "Introduction to R and RStudio"
subtitle: "Data Skills for Psychology — Module 1"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 1](https://psyteachr.github.io/ads-v3/01-intro.html).

## Today's Goals

By the end of this class, you will be able to:

-   Install R and RStudio
-   Install add-on packages
-   Get help for packages and functions
-   Create objects by writing and running code in the console

## What is R?

R is a **programming language** designed for statistics and data analysis.

. . .

Think of R as the **engine** of a car — it does all the work under the hood.

. . .

::: callout-tip
## Why R?

-   Free and open-source
-   Thousands of add-on packages
-   Massive community (especially in psychology!)
-   Makes your work **reproducible**
:::

## What is RStudio?

RStudio is an **Integrated Development Environment** (IDE) — it makes working with R much easier.

. . .

If R is the engine, RStudio is the **dashboard, steering wheel, and GPS** all in one.

. . .

::: callout-warning
Always launch **RStudio**, not R directly!
:::

# 🛠️ Setup Check {background-color="#16a085"}

## Activity: Installation Check

::: callout-tip
## Do this now!

1.  Open **RStudio** on your computer
2.  Check: do you see four panes?
3.  Give a thumbs up when you're ready!

If you haven't installed R and RStudio yet, follow the [installation guide](https://psyteachr.github.io/RSetGo/).
:::

# The RStudio IDE {background-color="#2c3e50"}

## The Four Panes

::::: columns
::: {.column width="50%"}
**Upper Left: Source**

-   Write and edit code
-   View data tables

**Lower Left: Console**

-   Type commands directly
-   See output
:::

::: {.column width="50%"}
**Upper Right: Environment**

-   See your objects
-   View your data

**Lower Right: Files/Plots/Help**

-   File browser
-   Plot viewer
-   Help documentation
-   Package manager
:::
:::::

## Activity: Explore the Panes

::: callout-tip
## Try it!

1.  Find the **Console** pane — type `1 + 1` and press Enter
2.  Find the **Environment** pane — is it empty?
3.  Find the **Help** pane — type `?mean` in the console
4.  Where did the help appear?
:::

. . .

::: callout-note
You can rearrange panes under **Tools \> Global Options \> Pane Layout**
:::

# ⚙️ RStudio Settings {background-color="#16a085"}

## Critical Settings to Change

You should **always start with a clean workspace** — no leftover objects from last time!

. . .

::: callout-tip
## Do this now!

Go to **Tools \> Global Options \> General**:

1.  [ ] **Uncheck** "Restore .RData into workspace at start-up"
2.  Set "Save workspace to .RData on exit" to **Never**
:::

. . .

::: callout-warning
You'll need to change these settings every time you update RStudio.
:::

# 📦 Packages {background-color="#2c3e50"}

## What are Packages?

R comes with built-in functions ("base R") — like the default apps on your phone.

. . .

**Packages** are add-ons created by the R community — like apps you download from the App Store.

. . .

There are thousands of packages for:

-   Data visualization 📊
-   Machine learning 🤖
-   Web scraping 🌐
-   Interactive dashboards 📱
-   Even playing Sudoku! 🧩

## Install vs. Load: A Key Distinction

::::: columns
::: {.column width="50%"}
### `install.packages()`

📲 Like **downloading** an app

-   Do it **once**
-   Only from the **console**
-   Stays until you remove it
:::

::: {.column width="50%"}
### `library()`

🚀 Like **launching** an app

-   Do it **every session**
-   Put it in your **script**
-   Only lasts until you close R
:::
:::::

## Activity: Install Your First Packages

::: callout-tip
## Run these in the console (not a script!)

```{r}
#| eval: false
#| echo: true
install.packages("tidyverse")
install.packages("beepr")
```
:::

. . .

::: callout-caution
## Important!

**Never** put `install.packages()` inside a script — only run it from the console. Why? You might accidentally reinstall and update a package, which could break your code.
:::

## Activity: Load and Use a Package

::: callout-tip
## Try it!

```{r}
#| eval: false
#| echo: true
# Load the package
library(beepr)

# Use a function from the package
beep()
```

🔊 Make sure your sound is on! What did you hear?
:::

. . .

::: callout-tip
## Now try this:

```{r}
#| eval: false
#| echo: true
beep(sound = 8)
```

What changed? Try values 1 through 11!
:::

# Functions {background-color="#2c3e50"}

## Anatomy of a Function

A **function** is a named piece of reusable code. It takes **arguments** (inputs) and produces output.

```{r}
#| eval: false
#| echo: true
function_name(argument1 = value1, argument2 = value2)
```

. . .

For example:

```{r}
#| eval: false
#| echo: true
beep(sound = 8)
#     ^         ^
#     |         |
#  argument   value
```

## The Tidyverse

`tidyverse` is a "meta-package" — a bundle of packages we'll use throughout this course:

| Package     | Purpose                 |
|-------------|-------------------------|
| `ggplot2`   | Data visualization      |
| `dplyr`     | Data manipulation       |
| `tidyr`     | Data tidying            |
| `readr`     | Reading data files      |
| `stringr`   | Working with text       |
| `forcats`   | Working with categories |
| `lubridate` | Working with dates      |

## Getting Help

There are several ways to get help with a function:

```{r}
#| eval: false
#| echo: true
# If the package is loaded
?beep

# If you're not sure about the package
??beep

# The full specification
?beepr::beep
```

. . .

Every help page has the same structure: **Description → Usage → Arguments → Details → Examples**

## Tab Auto-Complete

One of RStudio's best features! Start typing a function name and press **Tab**:

-   RStudio will show you matching functions
-   Select a function and press Tab again to see its arguments
-   Each argument comes with a brief description

. . .

::: callout-tip
## Try it!

Type `sam` in the console and press Tab. Select `sample(` and press Tab again to see the arguments.
:::

# Arguments In Practice {background-color="#2c3e50"}

## Example: `sample()`

```{r}
#| echo: true
# Randomly pick 5 letters
sample(x = letters, size = 5)
```

. . .

```{r}
#| echo: true
# Allow duplicates
sample(x = letters, size = 5, replace = TRUE)
```

## Named vs. Positional Arguments

These two lines do the **same thing**:

```{r}
#| eval: false
#| echo: true
# Named arguments (explicit)
sample(x = letters, size = 5, replace = TRUE)

# Positional arguments (implicit)
sample(letters, 5, TRUE)
```

. . .

With named arguments, **order doesn't matter**:

```{r}
#| eval: false
#| echo: true
sample(size = 5, replace = TRUE, x = letters)
```

. . .

::: callout-note
When learning, use named arguments — they make your code easier to read!
:::

# Objects {background-color="#2c3e50"}

## Creating Objects

Objects **store things** — numbers, text, dates, results. Use `<-` to assign values:

```{r}
#| echo: true
name <- "Maureen"
age <- 42
today <- Sys.Date()
halloween <- as.Date("2026-10-31")
```

. . .

To see what's inside an object, type its name:

```{r}
#| echo: true
name
today
```

## Types of Data

| Object | Type | Key Rule |
|------------------------|------------------------|------------------------|
| `name <- "Maureen"` | Character (text) | Must use quotes `" "` |
| `age <- 42` | Numeric (number) | No quotes |
| `today <- Sys.Date()` | Date (dynamic) | Changes each day! |
| `halloween <- as.Date("2026-10-31")` | Date (fixed) | Wrapped in `as.Date()` |

. . .

Check the **Environment pane** — your objects should appear there!

## Objects Can Interact

```{r}
#| echo: true
name <- "Maureen"
age <- 42
today <- Sys.Date()
halloween <- as.Date("2026-10-31")

decade <- age + 10
full_name <- paste(name, "Ritchey")
how_long <- halloween - today
```

. . .

```{r}
#| echo: true
decade
full_name
how_long
```

## Activity: Create Your Own Objects

::: callout-tip
## Type this into your console

1.  Create objects for **your** name, age, and a date you're looking forward to
2.  Calculate how many days until that date
3.  Use `paste()` to create a sentence like: *"Hi \[name\] — there are \[X\] days until \[event\]!"*

```{r}
#| eval: false
#| echo: true
# Example structure
name <- "Your Name"
today <- Sys.Date()
my_date <- as.Date("2026-12-25")
days_left <- my_date - today
message <- paste("Hi", name, "— there are",
                 days_left, "days to go!")
message
```
:::

# Getting Help {background-color="#2c3e50"}

## Your Help Toolkit

| Resource                    | When to Use                     |
|-----------------------------|---------------------------------|
| `?function_name`            | Know the function, need details |
| `??search_term`             | Not sure which function to use  |
| Google + "R" or "tidyverse" | Stuck on a concept              |
| AI tools (responsibly!)     | Need examples or explanations   |
| Package vignettes           | Want a tutorial-style guide     |
| Your classmates!            | Always a good idea              |

## A Note on AI

AI tools can help with coding — but be careful while you're building foundational skills.

. . .

> Using AI to do your exercises is like getting a forklift to lift weights at the gym for you. It defeats the purpose!

. . .

Build your skills first, then use AI to **accelerate** your work — not replace your learning.

# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   **R** is the language; **RStudio** is the tool that makes it easier
-   Always start with a **clean workspace**
-   `install.packages()` = download once; `library()` = load each session
-   **Functions** take arguments; use `?` to learn about them
-   **Objects** store your data; use `<-` to create them
-   When stuck: restart R, check help, Google it, ask a friend!

## Before Next Time

📖 **Read**: Chapter 1 of *Applied Data Skills* to lock in what we learned today

❓ **Questions?**
