---
title: "Data Visualization"
subtitle: "Data Skills for Psychology — Module 4"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 3](https://psyteachr.github.io/ads-v3/03-viz.html).

## Today's Goals

By the end of this class, you will be able to:

-   Describe the **layered grammar of graphics** behind `{ggplot2}`
-   Build a plot layer by layer: data, mapping, geoms, scales, themes
-   Pick an appropriate plot type for categorical vs. continuous data
-   Customize axes, labels, colours, and themes
-   Make your plots easier to read using jitter and facets
-   Save plots with `ggsave()`

## Setup

Make sure you have these packages loaded:

```{r eval=TRUE}
library(tidyverse)   # readr, dplyr, ggplot2, etc.
library(patchwork)   # combine plots side-by-side
library(ggthemes)    # extra themes
```

::: callout-tip
If you haven't installed `patchwork` or `ggthemes` yet, run in the **Console**:\
`install.packages(c("patchwork", "ggthemes"))`

You only need to do this once!
:::

## Our Dataset

We'll use data from the World Bank's [Poverty and Inequality Platform](https://pip.worldbank.org/poverty-calculator). Load it from the web:

```{r eval=TRUE}
pip <- read_csv("https://raw.githubusercontent.com/PsyTeachR/ads-v3/refs/heads/master/data/pip_gini_welfare_2010.csv")
```

. . .

Each row is one country in 2010. Key columns:

-   `region_name`, `region_code` — world region
-   `country_name`, `country_code` — country
-   `gini` — inequality (0 = perfect equality, 1 = complete inequality)
-   `welfare_mean` — mean welfare per capita
-   `welfare_median` — median welfare per capita

# The Grammar of Graphics {background-color="#2c3e50"}

## Plots are Built in Layers

`{ggplot2}` uses a **layered grammar of graphics**: plots are built up one piece at a time.

. . .

Think of it like building a presentation slide — each layer sits on top of the last:

1.  A blank canvas
2.  Variables mapped to **axes**
3.  A **geom** (the shape that represents your data)
4.  Maybe more geoms on top
5.  Axis tweaks and labels
6.  A **theme**

. . .

Each layer is independently customizable — and you **add** them with `+`.

## Set a Default Theme

The book uses the black-and-white theme, not the default grey one. Add this to your setup chunk:

```{r}
theme_set(theme_bw())   # applies to every plot you make
```

. . .

You can always override it on a per-plot basis — but this keeps things consistent.

# Building a Plot, Layer by Layer {background-color="#2c3e50"}

## Layer 1: A Blank Plot

Every plot starts with `ggplot()` and a data table:

```{r}
ggplot(data = pip)
```

. . .

That's it — a blank grey box. No variables yet, so there's nothing to draw.

## Layer 2: Mapping Variables

Use `aes()` (aesthetic mapping) to tell ggplot **which columns go where**:

```{r}
ggplot(data = pip,
       mapping = aes(x = gini, y = welfare_mean))
```

. . .

Now the axes exist, but there's still nothing *drawn* — we haven't told it **how** to represent the data.

. . .

::: callout-note
In practice, almost everyone drops the argument names and writes the code more like this:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean))
```
:::

## Layer 3: Add a Geom

**Geoms** are the shapes that represent your data — their functions all start with `geom_`. Add one with `+`:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point()
```

. . .

::: callout-warning
## Watch the `+`!

The `+` has to be at the **end** of the previous line, not at the start of the next one. If you write `+ geom_point()` on its own line, you'll get an error.
:::

## Multiple Geoms

You can stack geoms — order matters (later layers draw on top):

```{r}
# boxplot with points on top
ggplot(pip, aes(x = region_name, y = welfare_mean)) +
  geom_boxplot(width = .2) +
  geom_jitter(width = .2) +
  coord_flip()
```

. . .

```{r}
# points behind the boxplot
ggplot(pip, aes(x = region_name, y = welfare_mean)) +
  geom_jitter(width = .2) +
  geom_boxplot(width = .2) +
  coord_flip()
```

## Activity: Your First Plot

::: callout-tip
## Try it!

1.  Load the `pip` data with `read_csv()` (see the Setup slide).
2.  Run `theme_set(theme_bw())`.
3.  Make a scatterplot of `gini` on x and `welfare_mean` on y. Which geom should you use?
4.  **Bonus**: Add a second geom — `geom_smooth(method = "lm", formula = y ~ x)` — to show a trendline.
:::

## Common Geoms

There are a lot of different geoms, but these are ones that you'll see most in this class:

-   `geom_point`
-   `geom_jitter`
-   `geom_boxplot`
-   `geom_line`
-   `geom_histogram`
-   `geom_errorbar`
-   `geom_smooth`

# Styling Geoms {background-color="#2c3e50"}

## Styling the Whole Geom

To change how **all** points (or lines, or bars) look, set arguments **inside the geom** — not inside `aes()`:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(colour = "dodgerblue",
             alpha = 0.2,      # 20% transparency
             shape = 18,       # solid diamond
             size = 2) +
  geom_smooth(method = lm,
              formula = y ~ x,
              colour = "navy",
              linetype = 3)    # dotted line
```

## Style by Category

When you want the style to **represent a variable**, put it inside `aes()` instead:

```{r}
# colour points by region
ggplot(pip, aes(x = gini, y = welfare_mean,
                colour = region_code)) +
  geom_point(alpha = 0.6)
```

. . .

::: callout-warning
## The golden rule

-   **Inside `aes()`** → the aesthetic represents data
-   **Outside `aes()`** (in the geom) → the aesthetic is a fixed style
:::

## A note on `ggplot` syntax

If you find yourself getting tripped up by the syntax for `ggplot`, don't worry. This is normal!

It is common for researchers to find examples of what they want online, and then tweak the examples to fit their project.

# Picking the Right Plot {background-color="#2c3e50"}

## Match the Plot to the Data

Different kinds of data call for different plots. We'll walk through the most common cases:

-   **Counts of categories** → bar / column plot
-   **One continuous variable** → histogram
-   **Continuous variable by group** → violin, boxplot, combo
-   **Two continuous variables** → scatterplot + trendline

. . .

The [ggplot2 cheat sheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html) is a great quick reference.

## Counting Categories: `geom_bar()`

`geom_bar()` counts rows per category — only needs an `x` aesthetic:

```{r eval=TRUE}
ggplot(pip, aes(x = region_code)) +
  geom_bar()
```

. . .

Add `fill` inside `aes()` to colour bars by category:

```{r}
ggplot(pip, aes(x = region_code, fill = region_code)) +
  geom_bar()
```

## `geom_col()` for Pre-computed Counts

If your data **already has a count column**, use `geom_col()` (which needs both `x` and `y`):

```{r eval=TRUE}
count_data <- count(pip, region_code, sort = TRUE)

ggplot(count_data, aes(x = region_code, y = n)) +
  geom_col()
```

. . .

Use `fct_reorder()` to sort bars by size:

```{r}
ggplot(count_data,
       aes(x = fct_reorder(region_code, n), y = n)) +
  geom_col() +
  labs(x = "Region")
```

# One Continuous Variable {background-color="#2c3e50"}

## Histograms
`geom_histogram()` splits a continuous variable into bins and counts each bin:

```{r eval=TRUE}
ggplot(pip, aes(x = welfare_mean)) +
  geom_histogram(binwidth = 5, boundary = 0,
                 fill = "white", color = "black")
```

. . .

Always set `binwidth` or `bins` — the default usually isn't what you want.

## Grouped Continuous: Violin Plot

Violin plots show the shape of a distribution for each group:

```{r eval=TRUE}
ggplot(pip, aes(x = region_code, y = welfare_mean)) +
  geom_violin(scale = "width")
```


## Grouped Continuous: Boxplot

Boxplots show **summary statistics**, not shape:

```{r eval=TRUE}
ggplot(pip, aes(x = region_code, y = welfare_mean)) +
  geom_boxplot()
```


## What's in a boxplot:

-   The **middle line** is the median (50th percentile)
-   The **box** goes from the 25th to 75th percentile
-   The **whiskers** extend 1.5× the box height
-   The **points** are outliers

## Combo: Violin + Boxplot + Mean

You can layer geoms to show shape, summary, and mean all at once:

```{r}
ggplot(pip, aes(x = region_code, y = welfare_mean,
                fill = region_code)) +
  geom_violin(alpha = 0.4, scale = "width") +
  geom_boxplot(width = 0.25, fill = "white",
               alpha = 0.75, outlier.alpha = 0) +
  stat_summary(fun = mean, geom = "point",
               size = 2, shape = 18) +
  guides(fill = "none") +
  labs(x = "Region", y = "Mean Welfare per Capita")
```

## Activity: Compare Distributions

::: callout-tip
## Try it!

Using the `pip` data:

1.  Make a **histogram** of `gini` (binwidth around 0.02 is a reasonable start).
2.  Make a **violin plot** of `gini` by `region_code`.
3.  Make a **boxplot** of the same thing.
4.  **Bonus**: Which plot best shows that some regions have narrow inequality distributions while others are broad?
:::

# Two Continuous Variables {background-color="#2c3e50"}

## Scatterplot

`geom_point()` is the workhorse for two continuous variables:

```{r eval=TRUE}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point()
```

. . .

A good rule of thumb: put the **cause** (or predictor) on the x-axis and the **effect** (or outcome) on the y-axis.

## Adding a Trendline

`geom_smooth()` fits a line through the data. The shaded ribbon is the confidence interval.

```{r eval=TRUE}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = lm, formula = y ~ x)   # linear
```

## Other trendline options

-   `method = loess` — flexible, wavy line (default for small data)
-   `method = lm, formula = y ~ x + I(x^2)` — quadratic (one bend)
-   Add `se = FALSE` to hide the confidence ribbon

# Axes and Scales {background-color="#2c3e50"}

## The `scale_` Family

`scale_` functions control how each aesthetic maps to the plot. Match the function to the data type:

| Data type   | x-axis                 | y-axis                 |
|-------------|------------------------|------------------------|
| Continuous  | `scale_x_continuous()` | `scale_y_continuous()` |
| Categorical | `scale_x_discrete()`   | `scale_y_discrete()`   |

## Axis Limits

To zoom in without dropping data, use `coord_cartesian()`:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(alpha = 0.2) +
  coord_cartesian(ylim = c(0, 100))
```

. . .

::: callout-warning
Setting `limits` **inside `scale_`** actually drops data outside the range. This matters for things like boxplots and violins! Use `coord_cartesian()` to just zoom.
:::

# Themes and Labels {background-color="#2c3e50"}

## Built-in Themes

`{ggplot2}` ships with several themes; `{ggthemes}` adds many more:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(alpha = 0.2) +
  theme_light(base_size = 11)   # font size in points
```

. . .

Some favourites to try: `theme_minimal()`, `theme_bw()`, `theme_classic()`, `ggthemes::theme_gdocs()`.

## Theme Tweaks

`theme()` lets you override any theme element. Use `element_blank()` to remove something:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(alpha = 0.2) +
  theme_light(base_size = 11) +
  theme(plot.background = element_blank())   # remove outer border
```

. . .

Check `?theme` to see the (large!) list of things you can tweak.

## Labels with `labs()`

`labs()` adds a title, subtitle, caption, and axis labels all at once:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(alpha = 0.2) +
  labs(title = "Inequality vs. welfare across regions",
       subtitle = "Data from 2010",
       x = "Gini Index",
       y = "Mean Welfare per Capita",
       caption = "Source: World Bank PIP")
```

## Activity: Style a Plot

::: callout-tip
## Try it!

Starting from your scatterplot:

1.  Add `alpha = 0.3` to `geom_point()` so the points are transparent.
2.  Use `scale_x_continuous()` and `scale_y_continuous()` to add nicer axis labels.
3.  Apply a theme you like (try `theme_minimal()` or `theme_light()`).
4.  Use `labs()` to add a title, subtitle, and caption.
5.  **Bonus**: Colour the points by `region_code`.
:::

# Saving Plots {background-color="#2c3e50"}

## `ggsave()`

`ggsave()` saves the **last plot** you made by default:

```{r}
ggsave("my_plot.png", width = 5, height = 7)
```

. . .

Or save a plot you stored in an object:

```{r}
box <- ggplot(pip, aes(x = region_name, y = gini)) +
  geom_boxplot()

ggsave("box_plot.jpg", plot = box, width = 7, height = 7)
```

. . .

The file type comes from the extension (`.png`, `.jpg`, `.pdf`, `.svg`, …). Default units are inches, but you can pass `units = "cm"` or `"mm"`.

# Dealing with Overplotting {background-color="#2c3e50"}

## The Problem

When lots of points pile up on the same spot, you can't see what's there:

```{r}
# imagine simdat with 10,000 rows and ratings of 1–5
ggplot(simdat, aes(x = score, y = rating)) +
  geom_point()
```

. . .

Every rating is only 1, 2, 3, 4, or 5 — so all points squash onto 5 horizontal lines.

## Fix 1: Jitter

`geom_jitter()` nudges each point a random tiny amount:

```{r}
ggplot(simdat, aes(x = score, y = rating)) +
  geom_jitter(height = 0.2, width = 0, alpha = 0.1)
```

. . .

-   `height` / `width` — how far to nudge (in data units)
-   `alpha` — transparency, so overlapping points add up visibly

## Fix 2: Facets

`facet_wrap()` breaks the plot into small multiples — one panel per category:

```{r}
ggplot(simdat, aes(x = score)) +
  geom_histogram(binwidth = 5, boundary = 0,
                 fill = "dodgerblue", color = "black") +
  facet_wrap(~ rating, ncol = 1, labeller = label_both)
```

. . .

Read `~ rating` as **"by rating"**. `label_both` adds both the variable name and its value to each panel.

# Putting It All Together {background-color="#2c3e50"}

## A Publication-Ready Plot {.smaller}

Start simple, add layers until you're happy:

```{r}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(colour = "dodgerblue", alpha = 0.3) +
  geom_smooth(method = lm, formula = y ~ x,
              colour = "navy", se = FALSE) +
  scale_x_continuous(name = "Gini Index",
                     breaks = seq(0, 1, .1)) +
  scale_y_continuous(name = "Mean Welfare per Capita",
                     breaks = seq(0, 100, 20)) +
  coord_cartesian(xlim = c(.2, .7), ylim = c(0, 100)) +
  theme_light(base_size = 12) +
  theme(plot.background = element_blank()) +
  labs(title = "Higher inequality, lower welfare",
       subtitle = "Countries, 2010")
```

. . .

Build these plots **one layer at a time** — render at each step. It's much easier to find the broken layer that way.

## A Publication-Ready Plot
```{r eval=TRUE, echo=FALSE}
ggplot(pip, aes(x = gini, y = welfare_mean)) +
  geom_point(colour = "dodgerblue", alpha = 0.3) +
  geom_smooth(method = lm, formula = y ~ x,
              colour = "navy", se = FALSE) +
  scale_x_continuous(name = "Gini Index",
                     breaks = seq(0, 1, .1)) +
  scale_y_continuous(name = "Mean Welfare per Capita",
                     breaks = seq(0, 100, 20)) +
  coord_cartesian(xlim = c(.2, .7), ylim = c(0, 100)) +
  theme_light(base_size = 12) +
  theme(plot.background = element_blank()) +
  labs(title = "Higher inequality, lower welfare",
       subtitle = "Countries, 2010")
```

# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   **`ggplot() + aes() + geom_*()`** — the three minimum ingredients
-   **Inside `aes()`** → represents data; **outside `aes()`** → fixed style
-   **`geom_bar()`** counts rows; **`geom_col()`** uses a pre-computed count
-   **Histograms** show one continuous variable; **violins + boxplots** compare groups
-   **Scatterplots** + **`geom_smooth()`** for two continuous variables
-   **`facet_wrap(~ var)`** breaks a plot into small multiples
-   Use **`scale_` + `coord_cartesian` + `theme_*` + `labs`** to polish
-   Save with **`ggsave()`**

## Before Next Time

📖 **Read**: Chapter 3 of *Applied Data Skills* to solidify today's material

✍️ **Assignment**: TBD

❓ **Questions?**
