---
title: "Data Summaries"
subtitle: "Data Skills for Psychology — Module 5"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 4](https://psyteachr.github.io/ads-v3/04-summary.html) and [Appendix D: Data Types](https://psyteachr.github.io/ads-v3/app-datatypes.html).

## Today's Goals

By the end of this class, you will be able to:

-   Identify R's basic data types (numeric, character, logical, factor)
-   Understand vectors, lists, and tables as containers for data
-   Import data from CSV files
-   Explore and clean a dataset
-   Summarize data by groups
-   Use pipes to chain functions together
-   Count, group, and calculate percentages

## Setup

Make sure you have these packages loaded:

```{r}
library(tidyverse)   # readr, dplyr, ggplot2, etc.
library(janitor)     # clean_names()
```

::: callout-tip
You will probably need to install the **janitor** package before you try to load it. In the **Console**, run:\
`install.packages(janitor)`

You will only need to do this once!
:::

# Data Types {background-color="#2c3e50"}

## Why Data Types Matter

Every value in R has a **type**. Knowing the type tells you what you can do with it.

. . .

You can't take the average of someone's name, and you can't capitalize a number — the type determines what operations make sense.

. . .

The three basic types: **numeric**, **character**, and **logical**.

## Numeric Data

Numbers are stored as either **doubles** (decimals) or **integers** (whole numbers):

```{r}
typeof(10)    # double (the default)
typeof(10.0)  # also double
typeof(10L)   # integer (the L suffix forces it)
```

. . .

In practice, the double vs. integer distinction rarely matters. What matters is: *is it a number?*

```{r}
is.numeric(10)
is.numeric("10")   # quotes make it text, not a number!
```

## Character Data

**Characters** (also called "strings") are text wrapped in quotes:

```{r}
typeof("hello")
typeof("123")       # still a character — it has quotes!
```

. . .

Single or double quotes both work:

```{r}
greeting <- "Hello, world"
greeting <- 'Hello, world'   # same thing
```

. . .

::: callout-warning
`"10"` and `10` are **not** the same thing. One is text, the other is a number.
:::

## Logical Data

**Logical** values are `TRUE` or `FALSE` — the result of yes/no questions:

```{r}
class(TRUE)
class(FALSE)
```

. . .

They arise naturally from comparisons:

```{r}
10 > 5
10 == 5
```

. . .

::: callout-note
You might see `T` and `F` as shortcuts, but always spell out `TRUE` and `FALSE` to avoid confusion.
:::

## Checking Types

Two useful functions for inspecting types:

| Function         | What it tells you                |
|------------------|----------------------------------|
| `typeof()`       | The low-level storage type       |
| `class()`        | The high-level category          |
| `is.numeric()`   | Is it a number? (`TRUE`/`FALSE`) |
| `is.character()` | Is it text?                      |
| `is.logical()`   | Is it TRUE/FALSE?                |

. . .

For most purposes, `class()` is the one you'll use most often.

## Activity: What Type Is It?

::: callout-tip
## Predict, then check!

What type will R report for each of these? Use `typeof()` or `class()` to check.

```{r}
#| eval: false
100
100L
"100"
TRUE
"TRUE"
1 == 2
```
:::

## Factors

A **factor** is a special type for categorical data. It stores a fixed set of possible values (called **levels**) in a specific order.

```{r}
seasons <- factor("Spring",
                  levels = c("Spring", "Summer",
                             "Fall", "Winter"))
seasons
```

. . .

Why use factors? Because they control the **order** that categories appear in plots and tables.

. . .

We'll use factors when we import data later today, to control how delivery types and age groups are ordered.

## Dates and Times

Dates and times are stored as numbers with a special class:

```{r}
today <- as.Date("2025-01-15")
class(today)     # Date
```

. . .

R stores dates as the number of days since January 1, 1970. The `class` tells R to display them as readable dates instead of raw numbers.

. . .

We won't work much with dates today, but it's good to know they exist!

# Containers {background-color="#2c3e50"}

## Vectors: One Type, Many Values

A **vector** holds multiple values of the **same type**. Create one with `c()`:

```{r}
ages <- c(19, 34, 7, 67)
names <- c("Alice", "Bob", "Carol", "Dave")
```

. . .

What happens if you mix types?

```{r}
mixed <- c(1, "two", TRUE)
mixed    # everything becomes character!
```

. . .

::: callout-warning
R will silently **coerce** mixed types to the most flexible type (usually character). This can cause sneaky bugs!
:::

## Selecting from Vectors

Use square brackets `[]` to pull out elements by position:

```{r}
ages[1]          # first element
ages[c(2, 4)]   # second and fourth
ages[-3]         # everything except the third
```

. . .

You can also use logical vectors to filter:

```{r}
adults <- ages >= 18
ages[adults]
```

## Lists: Mixed Types Allowed

A **list** can hold different types of data in one container:

```{r}
participant <- list(
  id = 101,
  name = "Maureen",
  scores = c(85, 92, 78),
  consent = TRUE
)
```

. . .

Access items with `$` or double brackets `[[]]`:

```{r}
participant$name
participant[["scores"]]
```

. . .

::: callout-tip
We are primarily going to use a special kind of list: **tibbles** (a type of data frame).
:::

## Tables (Tibbles)

A **table** (or tibble) is the most important container for data analysis. It's a list of columns where every column has the same length:

```{r}
library(tidyverse)

survey <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Carol"),
  score = c(85, 92, 78)
)
survey
```

. . .

Useful functions for inspecting tables:

```{r}
nrow(survey)     # number of rows
ncol(survey)     # number of columns
glimpse(survey)  # quick overview
```

## Activity: Build a Tibble

::: callout-tip
## Try it!

1.  Create a tibble called `pets` with three columns: `name` (character), `species` (character), and `age` (numeric). Include at least 3 rows.
2.  Use `glimpse()` to check the column types.
3.  Use `$` to extract just the `name` column.
4.  **Bonus**: What class is the column you extracted?
:::

# Importing Data {background-color="#2c3e50"}

## Reading a CSV File

Often when we work with data, we want to load it from a separate file, such as a csv (comma-separated value) file.

Use `read_csv()` from the tidyverse to load data:

```{r}
# The csv file must be in a folder called "data" that lives in the same folder as this Quarto file.

births <- read_csv("data/12.1_delivery.csv")
```

. . .

Always check the loading summary! Look for:

-   Correct number of rows and columns
-   Expected data types (chr, dbl, etc.)
-   Any parsing problems

## Exploring the Import

```{r}
spec(births)     # column types in detail
glimpse(births)  # quick overview of all columns
```

. . .

Do you remember the other methods we have for exploring data?

## Activity: Load and Explore

::: callout-tip
## Try it!

1.  Load the births dataset
2.  Run `spec()` and `glimpse()` — how many rows and columns?
3.  Are there any columns that look useless?
:::

## Skipping Columns on Import

If a column has no useful info, skip it at load time using `col_skip`:

```{r}
# cols refers to "columns"
# Here we are telling R about the data columns BEFORE we load the data
birth_cols <- cols(
  SIMDQuintileQF = col_skip()
)

# Now reload
births <- read_csv("data/12.1_delivery.csv",
                   col_types = birth_cols)
```

. . .

This is cleaner than loading everything and dropping columns later.

## Setting Factor Levels on Import

You can control the order of categorical variables right from the start, by defining them as a **factor** and setting their **levels**:

```{r}
# Again, here we're telling R about the data columns BEFORE we load the data
birth_cols <- cols(
  SIMDQuintileQF = col_skip(),
  AgeGroup = col_factor(
    levels = c("Under 25", "25-34",
               "35 and over", "Unknown")),
  Induced = col_factor(
    levels = c("Not Induced", "Induced", "Unknown"))
)

# Now reload
births <- read_csv("data/12.1_delivery.csv",
                   col_types = birth_cols)
```

. . .

Why bother? Because plots and tables will display groups in **this** order.

## Cleaning Column Names

Column names with mixed case are annoying. The `janitor` package fixes this:

```{r}
births <- clean_names(births)
colnames(births)
```

. . .

`clean_names()` converts everything to lowercase with underscores: `AgeGroup` becomes `age_group`.

. . .

::: callout-tip
## Good habit

Run `clean_names()` every time you import a new dataset!
:::

# The `$` Operator and Pipes {background-color="#2c3e50"}

## Selecting Columns with `$`

The `$` lets you pull out a single column from a tibble or data frame:

```{r}
years <- births$financial_year
years
```

. . .

You can also grab specific rows with square brackets:

```{r}
births$financial_year[1]
births$livebirths[c(20, 30, 40)]
```

## Pipes: The Big Idea {.smaller}

### Without pipes — lots of intermediate objects:

```{r}
c_sections <- filter(births,
                     delivery == "Caesarean - Emergency")
c_per_year <- summarise(c_sections,
                        n = sum(livebirths),
                        .by = financial_year)
c_sorted <- arrange(c_per_year, desc(n))
```

. . .

### With pipes — one clean chain:

```{r}
c_per_year <- births |>
  filter(delivery == "Caesarean - Emergency") |>
  summarise(n = sum(livebirths),
            .by = financial_year) |>
  arrange(desc(n))
```

## How to Read a Pipe

Translate `|>` as **"and then"**:

```{r}
births |>                                     # start with births AND THEN
  filter(delivery == "Caesarean - Emergency") |>  # keep only emergency c-sections AND THEN
  summarise(n = sum(livebirths),                  # sum up livebirths AND THEN
            .by = financial_year) |>
  arrange(desc(n))                                # sort largest to smallest
```

. . .

The pipe sends the result of each step as the **first argument** to the next function.

. . .

::: callout-note
You'll see two pipes in the wild: `|>` (base R, newer) and `%>%` (tidyverse/magrittr, older). They work almost identically. We'll use `|>`.
:::

## Activity: Practice with Pipes

::: callout-tip
## Rewrite this using pipes

```{r}
spontaneous <- filter(births, delivery == "Spontaneous")
spont_by_age <- summarise(spontaneous,
                          n = sum(livebirths),
                          .by = age_group)
spont_sorted <- arrange(spont_by_age, desc(n))
```

Try rewriting this as a single piped chain!

For now, don't worry too much about what each step does - pay attention to the way that the variables are passed from one line to the next.
:::

# Counting {background-color="#2c3e50"}

## The `count()` Function

`count()` tallies up rows per group:

```{r}
count(births, delivery)
```

. . .

Add `sort = TRUE` to see the most common first:

```{r}
count(births, induced, sort = TRUE)
```

## Counting Multiple Variables

You can count combinations of columns:

```{r}
count(births, age_group, induced, sort = TRUE)
```

. . .

::: callout-note
`count()` counts **rows**, not births! The `livebirths` column contains the actual numbers per row. To get total births, you need a different function that we'll go over shortly: `summarise()` with `sum()`.
:::

## Adding Percentages

Use `mutate()` to add a percentage column:

```{r}
count(births, delivery) |>
  mutate(percent = n / sum(n),
         percent = round(100 * percent))
```

. . .

Breaking it into two steps (divide, then round) is cleaner than nesting: `round(100 * n / sum(n))`. But either is technically correct.

## Activity: Counting Practice

::: callout-tip
## Try these

1.  Count the number of rows per `age_group`
2.  Count the number of rows by `delivery` and `age_group` together
3.  Add a percentage column to your delivery counts

Which delivery type is most common in the data?
:::

# Summarize {background-color="#2c3e50"}

## The `summarize()` Function

`summarize()` (or sometimes, `summarise()`) computes summary statistics and returns a new, smaller table:

```{r}
summarize(births,
          total_births = sum(livebirths))
```

. . .

Common summary functions:

| Function         | What it does  |
|------------------|---------------|
| `n()`            | Count rows    |
| `sum()`          | Add up values |
| `mean()`         | Average       |
| `median()`       | Middle value  |
| `min()`, `max()` | Extremes      |
| `quantile()`     | Percentiles   |

## A Note on Missing Values

If you get `NA` from a summary function, it usually means missing data:

```{r}
values <- c(1, 2, 4, 3, NA, 2)
mean(values)              # returns NA!
mean(values, na.rm = TRUE) # removes NAs first
```

. . .

This is logical: the average of 1 + 2 + "I don't know" isn't 1.5 — it's "I don't know."

# Grouping {background-color="#2c3e50"}

## `group_by()` + `summarize()`

You can group the data before you summarize it:

```{r}
births |>
  group_by(delivery) |>
  summarize(total = sum(livebirths)) |>
  ungroup()
```

. . .

::: callout-warning
Always call `ungroup()` when you're done! Forgetting this causes mysterious bugs later.
:::

## Activity: Summarize the Data

::: callout-tip
## Answer these questions with code

1.  What is the total number of live births across the whole dataset?
2.  What is the total number of live births per age group?
3.  What is the mean number of live births per row, broken down by delivery type?
:::

## Sorting with `arrange()`

`arrange()` sorts your results:

```{r}
births |>
  group_by(delivery) |>
  summarise(total = sum(livebirths)) |>
  ungroup() |>
  arrange(desc(total))
```

. . .

Use `desc()` for descending order (largest first).

## Multiple Groupings

Group by multiple variables:

```{r}
births |>
  group_by(delivery, financial_year) |>
  summarise(n = sum(livebirths)) |>
  ungroup()
```

. . .

::: callout-tip
## Try it

Reverse the order of `delivery` and `financial_year` in `group_by()`. How does the output change?
:::

# Putting It All Together {background-color="#2c3e50"}

## A Full Analysis Pipeline {.smaller}

Combine everything into one clean chain:

```{r}
births_per_year_type_age <- births |>
  mutate(
    year = str_extract(financial_year, ".{4}"),
    year = as.integer(year)
  ) |>
  filter(age_group != "Unknown",
         delivery != "Not Known") |>
  summarise(
    n = sum(livebirths),
    .by = c(year, delivery, age_group)
  ) |>
  mutate(
    pcnt = n / sum(n) * 100,
    .by = c(year, age_group)
  )

head(births_per_year_type_age)
```

## Visualizing the Result

```{r}
births_per_year_type_age |>
  ggplot(aes(x = year, y = pcnt, color = age_group)) +
  geom_line() +
  facet_wrap(~delivery, scales = "free_y") +
  labs(x = NULL,
       y = "Percent of Live Births per Year",
       color = "Age Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

. . .

From raw data to publication-quality plot — all in one reproducible pipeline!

# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   **`read_csv()`** loads data; use `col_skip()` and `col_factor()` to clean on import
-   **`clean_names()`** standardizes column names (always do this!)
-   **`|>`** (pipe) chains functions together — read it as "and then"
-   **`count()`** tallies rows per group
-   **`summarise()`** computes stats like `sum()`, `mean()`, `n()`
-   **`group_by()`** computes summaries per group
-   **`mutate()`** adds or modifies columns (e.g., percentages)
-   Always **`ungroup()`** after `group_by()`!

## Before Next Time

📖 **Read**: Chapter 4 of *Applied Data Skills* to solidify today's material

✍️ **Assignment**: You will practice these skills using a real psychology dataset. See the website for more information.

❓ **Questions?**
