---
title: "Data Wrangling"
subtitle: "Data Skills for Psychology — Module 6"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 9: Data Wrangling](https://psyteachr.github.io/ads-v3/09-wrangle.html).

## Today's Goals

By the end of this class, you will be able to:

-   Select and filter data for relevance
-   Rename, reorder, and create new columns
-   Use `case_when()` to recode values with multiple conditions
-   Handle missing data thoughtfully (convert, ignore, or drop)
-   Chain wrangling steps together with the pipe (`|>`)

## Setup

Make sure you have these packages loaded:

```{r}
library(tidyverse)   # readr, dplyr, tidyr, stringr, etc.
library(readxl)      # read .xlsx files
library(janitor)     # clean_names(), round_half_up()
```

::: callout-tip
You can grab the [Data transformation cheat sheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html) — it's a handy reference for everything we'll cover today.
:::

## The Dataset

We'll use UN data on **public transport access** (SDG 11.2.1):

> *"Proportion of population that has convenient access to public transport, by sex, age and persons with disabilities."*

. . .

Download `SDG-11.2.1.xlsx` into your `data/` folder:

[psyteachr.github.io/ads-v3/data/SDG-11.2.1.xlsx](https://psyteachr.github.io/ads-v3/data/SDG-11.2.1.xlsx)

```{r}
transport <- readxl::read_xlsx("data/SDG-11.2.1.xlsx") |>
  janitor::clean_names()
```

# The dplyr Toolkit {background-color="#2c3e50"}

## Six Verbs Do Most of the Work

The <pkg>dplyr</pkg> package gives you a small set of verbs that combine to handle most wrangling tasks:

| Verb          | What it does                               |
|---------------|--------------------------------------------|
| `select()`    | pick **columns**                           |
| `filter()`    | pick **rows**                              |
| `arrange()`   | sort rows                                  |
| `mutate()`    | add or change columns                      |
| `summarise()` | collapse rows into summary statistics      |
| `group_by()`  | apply other verbs separately to each group |

. . .

You already met `summarise()` and `group_by()` before. Today we focus on the other four.

## Exploring First {background-color="#16a085"}

## Glimpse and Count

Before wrangling, look at what you have:

```{r}
glimpse(transport)
```

. . .

`count()` is your friend for categorical columns:

```{r}
count(transport, age, sex)
```

. . .

It looks like most rows aggregate across age and sex. If we want gender comparisons, we'll need to filter.

## Spotting Useless Columns

Some columns have only `NA`:

```{r}
count(transport, time_coverage)   # all 2473 rows are NA
```

. . .

Others have the same value in every row:

```{r}
count(transport, goal)            # all 2473 rows are 11
```

. . .

These won't help with analysis — good candidates to drop.

## Activity: Explore the Data

::: callout-tip
## Try it!

1.  Load the `transport` dataset
2.  Use `count()` to figure out how many **countries** are in the data
3.  Use `count()` to figure out how many different **years** are covered
4.  **Bonus:** Find one more column that has the same value in every row
:::

# Select {background-color="#2c3e50"}

## Picking Columns by Name

`select()` chooses which columns to keep:

```{r}
# select a single column
value_col <- transport |> select(value)

# select several columns, in a new order
country_year <- transport |>
  select(geo_area_name, time_period, value)
```

. . .

You can rename columns at the same time:

```{r}
country_year <- transport |>
  select(country = geo_area_name,
         year    = time_period,
         value)
```

## Selecting a Range

Use a colon to grab a span of consecutive columns:

```{r}
country_year <- transport |>
  select(geo_area_name:value)
```

. . .

::: callout-warning
The colon is fast, but check your output! The columns must already be in the order you want.
:::

## Un-selecting Columns

Put a minus sign in front to **drop** columns instead:

```{r}
transport_clean <- transport |>
  select(-time_coverage,
         -upper_bound,
         -lower_bound,
         -base_period)
```

. . .

For ranges, wrap in parentheses first:

```{r}
transport_clean <- transport |>
  select(-(time_coverage:base_period))
```

## Select Helpers

When you have many columns, name patterns are powerful:

| function        | what it picks                           |
|-----------------|-----------------------------------------|
| `everything()`  | all columns                             |
| `starts_with()` | columns whose names start with a string |
| `ends_with()`   | columns whose names end with a string   |
| `contains()`    | columns whose names contain a string    |
| `where()`       | columns where a function returns `TRUE` |
| `all_of()`      | all columns named in a vector           |
| `any_of()`      | any columns from a vector that exist    |

## Helpers in Action

```{r}
# all columns starting with "geo_"
transport |> select(starts_with("geo_"))

# all numeric columns
transport |> select(where(is.numeric))

# drop a vector of column names
one_val_cols <- c("goal", "target", "indicator",
                  "series_code", "time_coverage",
                  "upper_bound", "lower_bound",
                  "base_period", "geo_info_url",
                  "custom_breakdown", "reporting_type",
                  "units")

transport_clean <- transport |>
  select(-all_of(one_val_cols))
```

# Rename and Relocate {background-color="#2c3e50"}

## Cleaner Renames and Reorders

You *can* rename and reorder inside `select()`, but dedicated verbs make pipelines easier to read:

```{r}
transport_clean <- transport |>
  select(-all_of(one_val_cols)) |>
  rename(year    = time_period,
         country = geo_area_name) |>
  relocate(age, sex, .after = value) |>
  relocate(geo_area_code, .before = cities)
```

. . .

`.after` and `.before` take a column name and tell R where to put the moved columns.

# Filter {background-color="#2c3e50"}

## Filtering Rows

`filter()` keeps rows that meet a condition:

```{r}
# rows where country is Togo
transport_clean |> filter(country == "Togo")

# rows where country is NOT Togo
transport_clean |> filter(country != "Togo")

# rows where value is more than 99
transport_clean |> filter(value > 99)
```

. . .

::: callout-warning
Use `==` (double equals) to compare. A single `=` is for **assignment**.
:::

## Multiple Conditions

Separate conditions with commas — rows must match **all** of them:

```{r}
togo_2020 <- transport_clean |>
  filter(country == "Togo",
         year == 2020)
```

. . .

Use `|` for **or** (and `&` if you prefer being explicit about "and"):

```{r}
extreme_values <- transport_clean |>
  filter(value < 10 | value > 90)
```

## The `%in%` Operator

For "match any of these values," use `%in%` rather than chaining lots of `|`s:

```{r}
islands <- c("Solomon Islands", "Northern Mariana Islands")

only_islands <- transport_clean |>
  filter(country %in% islands)

# negate with ! before the variable
no_islands <- transport_clean |>
  filter(!country %in% islands)
```

## Comparison Operators

| Operator   | Meaning                      |
|------------|------------------------------|
| `A < B`    | A is less than B             |
| `A <= B`   | A is less than or equal to B |
| `A > B`    | A is greater than B          |
| `A >= B`   | A is greater than or equal   |
| `A == B`   | A exactly equals B           |
| `A != B`   | A does not equal B           |
| `A %in% B` | A is an element of vector B  |

## Filtering with Functions

You can pass any function that returns `TRUE`/`FALSE`. For text, `str_detect()` finds substrings:

```{r}
transport_clean |>
  filter(str_detect(country, "Island")) |>
  select(1:6)
```

. . .

::: callout-note
`str_detect()` is **case-sensitive**. Use `tolower()` or `toupper()` first if you need to match regardless of case.
:::

## Always Check Your Filter

::: callout-warning
`filter()` is powerful — and easy to misuse. With multiple criteria, it's surprisingly common to specify slightly the wrong thing.

**Always check your output.** Use `count()` or `glimpse()` afterwards to confirm you got what you intended.
:::

```{r}
transport_filtered <- transport_clean |>
  filter(sex %in% c("FEMALE", "MALE"),
         age == "ALLAGE")

# verify: should be only MALE/FEMALE and ALLAGE
count(transport_filtered, sex, age)
```

## Activity: Filter Practice

::: callout-tip
## Try it!

Starting from `transport_clean`:

1.  Filter to rows from the year **2015** only
2.  Filter to rows where the country name contains **"United"**
3.  Combine both: rows from 2015 where the country contains "United"
4.  **Bonus:** Use `count()` to verify each filter returned what you expected
:::

# Arrange {background-color="#2c3e50"}

## Sorting Rows

`arrange()` sorts your data — alphabetically for text, ascending for numbers:

```{r}
transport_arranged <- transport_filtered |>
  arrange(country, desc(year))
```

. . .

Wrap a column in `desc()` to reverse the order. You can sort by multiple columns — earlier columns take priority.

## Custom Sort Order with Factors

For non-alphabetical category orders, convert to a factor and set the levels:

```{r}
transport_factor <- transport_filtered |>
  mutate(sex = factor(sex, levels = c("MALE", "FEMALE")))

transport_factor |>
  filter(year %in% 2001:2003) |>
  select(country, year, sex) |>
  arrange(year, sex)
```

# Mutate {background-color="#2c3e50"}

## Adding and Changing Columns

`mutate()` creates a new column or overwrites an existing one. Syntax: `new_column = expression`.

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value)
  )
```

. . .

You can build several columns in one `mutate()`, separated by commas. Later definitions can use earlier ones.

## Mutate with Conditions

A Boolean expression returns `TRUE`/`FALSE`:

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    above_median = value > median(value)
  )
```

. . .

::: callout-warning
If you reuse a column name, you **overwrite** the old column. Make sure that's what you want!
:::

## `case_when()` for Multiple Categories

When you need more than two outcomes, use `case_when()`:

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    category = case_when(
      value < 70  ~ "low",
      value < 85  ~ "average",
      value >= 85 ~ "high",
      .default    = NA
    )
  )
```

. . .

R checks each condition top to bottom and assigns the first match.

## Combining Conditions in `case_when()`

You can mix multiple variables in each rule:

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    category = case_when(
      urban  & value < 75 ~ "low",
      urban  & value < 85 ~ "average",
      urban  & value >= 85 ~ "high",
      !urban & value < 20 ~ "low",
      !urban & value < 60 ~ "average",
      !urban & value >= 60 ~ "high",
      .default = NA
    )
  )
```

::: callout-tip
`.default` uses `=`, while every other rule uses `~`. Easy to mix up!
:::

## Activity: Build Your Own Categories

::: callout-tip
## Try it!

Using `transport_factor` (or your own version):

1.  Make `value` numeric with `as.numeric()`
2.  Add a column `era` that is:
    -   `"early"` if year is before 2010
    -   `"mid"` if year is 2010 to 2019
    -   `"recent"` if year is 2020 or later
3.  Use `count()` to see how many rows fall in each era
4.  **Bonus:** Add a `value_label` column that says `"high"` if value \> 80, otherwise `"not high"`
:::

# Summarise & Group {background-color="#2c3e50"}

## Quick Refresher: `summarise()`

`summarise()` collapses many rows into one (or one per group). Name each summary on the left:

```{r}
transport_data |>
  summarise(
    mean = mean(value, na.rm = TRUE),
    min  = min(value, na.rm = TRUE),
    max  = max(value, na.rm = TRUE)
  )
```

. . .

Without grouping, you get a single-row summary of the whole dataset.


# Complications {background-color="#2c3e50"}

## Rounding Surprises

```{r}
round(0.5)   # returns 0 (!)
round(1.5)   # returns 2
```

. . .

R rounds **0.5 to the nearest even number** — this prevents bias when many values are halfway. It's not a bug.

. . .

If you want school-style rounding (always up at .5), use <pkg>janitor</pkg>:

```{r}
round_half_up(0.5)   # returns 1
round_half_up(1.5)   # returns 2
```

# Missing Values {background-color="#2c3e50"}

## Why Missing Values Need Care

Best practice: leave missing cells **empty** in your raw data.

. . .

If "missing" appears as a word, the whole column becomes character data — you can't do math on it.

. . .

If 0 or 999 stand in for missing, those values may sneak into calculations as if they were real.

## Recoding "missing" as `NA`

Suppose someone wrote `"missing"` instead of leaving the cell blank:

```{r}
missing_bad <- transport_data |>
  mutate(value = case_when(
    year == 2000 & country == "Uzbekistan" ~ "missing",
    .default = as.character(value)
  ))
```

. . .

Now `mean(value)` errors out — it's a character column, not numeric.

## Fix It with `na_if()`

Convert the bad string to `NA`, then back to numeric:

```{r}
missing_data <- missing_bad |>
  mutate(
    value = na_if(value, "missing"),
    value = as.numeric(value)
  )
```

. . .

Try a summary now:

```{r}
missing_data |>
  summarise(
    mean = mean(value),
    .by  = country
  )
```

You'll get `NA` for any country with a missing value.

## Telling `mean()` to Ignore `NA`s

Most summary functions take `na.rm = TRUE`:

```{r}
missing_data |>
  summarise(
    mean = mean(value, na.rm = TRUE),
    .by  = country
  )
```

. . .

Why isn't this the default? Because `NA` means "I don't know," and the average of *known* values plus *unknown* values is *unknown*. R wants you to make the call deliberately.

## Counting Missing Values

`is.na()` returns `TRUE`/`FALSE` for each value. Combine with `sum()` or `mean()`:

```{r}
missing_data |>
  group_by(country) |>
  summarise(
    n_valid     = sum(!is.na(value)),
    n_missing   = sum(is.na(value)),
    prop_missing = mean(is.na(value))) |>
  ungroup()
```

## Dropping Rows with `drop_na()`

```{r}
# remove any row with any missing value
complete_data <- missing_data |> drop_na()

# remove only rows missing a specific column
complete_values <- missing_data |> drop_na(value)
```

. . .

::: callout-warning
`drop_na()` with no arguments drops a row if **any** column is missing — including columns you don't care about (notes, comments, etc.). Be specific about which columns matter.
:::

## Activity: Handle Missing Data

::: callout-tip
## Try it!

Using `missing_data` (or recreate it from the slides):

1.  Calculate the mean `value` overall, ignoring missing values
2.  Count how many rows are missing a `value`
3.  Create `complete_data` that drops only rows missing `value` — leave other columns alone
4.  **Bonus:** Use `summarise()` to find which country has the highest proportion of missing values
:::

# Applications to the Screentime dataset {background-color="#2c3e50"}

## Switching Datasets

Let's apply everything to the [`screentime`](../datasets/screentime.html) dataset — over 120,000 adolescents' digital screen time and mental well-being scores (Przybylski & Weinstein, 2017).

. . .

Key columns we'll use:

-   `mwbi` — mental well-being index (composite score)
-   `WBOptimf`, `WBUseful`, `WBRelxd`, ... — individual well-being items
-   `male`, `minority`, `deprived` — demographic indicators (1 = yes, 0 = no)
-   `REGION` — UK region

```{r}
screentime <- read_csv("data.csv")
```

## Select: Pick Useful Columns

Several ways to grab what you want:

```{r}
# specific columns by name
screentime |>
  select(REGION, male, minority, deprived, mwbi)

# a range with the colon
screentime |>
  select(REGION, male:mwbi)

# combine with a helper
screentime |>
  select(REGION, male:mwbi, starts_with("WB"))

# drop a single column
screentime |>
  select(-REGION)
```

## Filter: Keep Useful Rows

The `mwbi` column uses negative numbers as missing-value codes — filter them out:

```{r}
screentime_select <- screentime |>
  select(REGION, male:mwbi, starts_with("WB"))

# keep only valid well-being scores
screentime_select |>
  filter(mwbi > 0)

# stack conditions: valid scores AND a specific region
screentime_select |>
  filter(mwbi > 0,
         REGION == "East of England")
```

::: callout-warning
Always check how missing data is coded in your dataset. Here, "missing" looks like a real (negative) number — easy to miss!
:::

## Mutate: Build a New Composite

Average two well-being items into a new column:

```{r}
screentime_select |>
  filter(mwbi > 0) |>
  mutate(new_measure = (WBOptimf + WBUseful) / 2)

```

. . .

::: callout-warning
**Watch out!** `mean(c(WBOptimf, WBUseful))` would average across the *whole column* — giving every row the same value. For a per-row average, use arithmetic (`(a + b) / 2`) or `rowMeans()`.
:::

## Group and Summarise

Compare mean well-being across demographic groups:

```{r}
screentime_select |>
  filter(mwbi > 0) |>
  mutate(new_measure = (WBOptimf + WBUseful) / 2) |>
  group_by(male, minority) |>
  summarise(
    mean_new = mean(new_measure),
    n        = n()) |>
  ungroup()
```


## Putting It All Together

The whole workflow as one pipeline:

```{r}
screentime |>
  select(REGION, male:mwbi, starts_with("WB")) |>
  filter(mwbi > 0,
         REGION == "East of England") |>
  mutate(new_measure = (WBOptimf + WBUseful) / 2) |>
  group_by(male, minority) |>
  summarise(
    mean_new = mean(new_measure),
    n        = n()) |>
  ungroup()
```

. . .

Read it top-to-bottom: *take screentime, then select these columns, then filter to valid scores in this region, then make a new measure, then summarise by demographic group.*


# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   The dplyr verbs (`select`, `filter`, `arrange`, `mutate`, `summarise`, `group_by`) compose well — chain them with `|>`
-   **Always check your output** after each step (use `glimpse()`, `count()`, `head()`)
-   `case_when()` is your friend for multi-way recoding
-   Missing values aren't bugs — they're information. Decide how to handle them on purpose

