---
title: "Data Tidying"
subtitle: "Data Skills for Psychology — Module 7"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 8: Data Tidying](https://psyteachr.github.io/ads-v3/08-tidy.html).

## Today's Goals

By the end of this class, you will be able to:

-   Recognize what makes data "tidy" (and what makes it messy)
-   Reshape data between **wide** and **long** formats
-   Split a single column into multiple columns
-   Chain tidying steps together with the pipe (`|>`)

## Setup

Make sure these packages are loaded:

```{r}
library(tidyverse)   # readr, dplyr, tidyr, stringr, etc.
library(readxl)      # read .xlsx files
library(janitor)     # clean_names()
```

::: callout-tip
Grab the [Data tidying cheat sheet](https://rstudio.github.io/cheatsheets/html/tidyr.html) — it's a great reference for the `pivot_*` and `separate*()` functions.
:::

## Download Today's Data

Save these into your `data/` folder:

-   [tidy.csv](https://psyteachr.github.io/ads-v3/data/tidy.csv)
-   [untidy.csv](https://psyteachr.github.io/ads-v3/data/untidy.csv)

. . .

Both contain the same information — proportion of women in parliaments across countries and years — but stored very differently.

# Why Tidy? {background-color="#2c3e50"}

## A Reality Check

> "Every messy dataset is messy in its own way." — Hadley Wickham

. . .

Data tidying is often the **most time-consuming part** of an analysis.

. . .

Moving between formats that are easy for **humans to read** and formats that are easy for **computers to use** takes practice — that conceptual shift is the real skill of this chapter.

## Three Words to Pin Down

**Observation:** all the information about a single "thing" in a single condition (e.g., one country in one year).

. . .

**Variable:** one type of information about that observation (e.g., year, number of MPs).

. . .

**Value:** the data for one variable for one observation (e.g., `2025`).

## The Three Rules of Tidy Data

For data to be "tidy":

1.  Each **observation** has its own **row**
2.  Each **variable** has its own **column**
3.  Each **value** has its own **cell**

. . .

When data is tidy, you can keep using the same code patterns even when the dataset grows or changes shape.

# Untidy vs Tidy {background-color="#2c3e50"}

## An Untidy Example

```{r}
untidy <- read_csv("data/untidy.csv", show_col_types = FALSE)
untidy
```

. . .

What's "off" about this table?

-   The `country` column squeezes in two values (name and code)
-   Each year has its own column (`1950`, `1975`, `2000`, `2025`)
-   Each cell holds **two numbers**: women / members

## Why That Hurts

Suppose you want the average proportion of women per year. In Excel you would:

1.  Split each year column into `women_*` and `members_*`
2.  Compute proportion for each year (4 new columns)
3.  Average each new column separately

. . .

If a new year of data arrives, you redo every step. Painful.

## The Tidy Version

```{r}
tidy_data <- read_csv("data/tidy.csv", show_col_types = FALSE)
tidy_data
```

. . .

Each row is one country in one year. Five columns: `country_name`, `country_code`, `year`, `members`, `women`.

## And the Calculation Becomes…

```{r}
tidy_data |>
  mutate(prop_women = women / members) |>
  summarise(
    avg_prop_women = mean(prop_women),
    .by = year
  )
```

. . .

The same code works whether you have 4 years or 40. That's the power of tidy.

# Reshaping Data {background-color="#2c3e50"}

## Wide vs Long

**Wide format**: all measurements about one thing on the same row.

**Long format**: each measurement on its own row.

. . .

You'll often need to convert between them. Think of it as the R version of a pivot table in Excel.

## `pivot_longer()`: Wide → Long {background-color="#16a085"}

## The Idea

`pivot_longer()` takes several columns and stacks them into two new columns:

-   one for the **old column names**
-   one for the **values** that were in those columns

. . .

Three key arguments:

-   `cols` — which columns to "fold up"
-   `names_to` — name of the new column for the old headers
-   `values_to` — name of the new column for the old values

## In Code

```{r}
long <- pivot_longer(
  data = untidy,
  cols = `1950`:`2025`,
  names_to = "year",
  values_to = "women_members"
)

long
```

. . .

Now there's one row per country-year, with `year` as a real variable.

. . .

::: callout-warning
Because the column names are **numbers**, you must wrap them in backticks. Without them, R thinks `1950:2025` means "columns 1950 through 2025."

```         
Error in `pivot_longer()`:
! Can't select columns past the end.
ℹ Locations 1950, 1951, …, 2025 don't exist.
```
:::

## Activity: Sketch a Long Version

::: callout-tip
## Try it!

Convert this table to long format. You don't need to write code — just sketch it on paper or in a spreadsheet.

| band               | twitter | instagram |
|:-------------------|:--------|:----------|
| The Beatles        | 3.8     | 3.8       |
| The Rolling Stones | 3.4     | 3.1       |
| One Direction      | 31.3    | 22.8      |

1.  How many rows should the long version have?
2.  What should the new column names be?
3.  **Bonus:** What would the long version look like if we *also* had a `youtube` column?
:::

## `pivot_wider()`: Long → Wide {background-color="#16a085"}

## The Idea

`pivot_wider()` is the reverse — it spreads one column's **values** into new column **names**.

. . .

Three key arguments:

-   `id_cols` — column(s) that uniquely identify each new row
-   `names_from` — column whose values become the new headers
-   `values_from` — column whose values fill the new cells

## In Code

```{r}
wide <- pivot_wider(
  data = long,
  id_cols = year,
  names_from = country,
  values_from = women_members
)

wide
```

. . .

Now each row is a year, and each country has its own column. Useful for some kinds of comparison or display.

# One Value Per Cell {background-color="#2c3e50"}

## The Problem

Look back at our `long` data — the `women_members` column still holds **two values** in each cell, like `"3 / 257"`.

. . .

Tidy data needs **one value per cell**, so we need to split this column.

## `separate()` to the Rescue

```{r}
split_data <- separate(
  data = long,
  col = women_members,        # the column to split
  into = c("women", "members"),  # the new column names
  sep = " / ",                # how to split
  remove = TRUE,              # drop the original column?
  convert = TRUE              # try to fix data types?
)

split_data
```

. . .

`convert = TRUE` is handy — it'll turn `"3"` and `"257"` into actual numbers automatically.

## Activity: Split the `country` Column

::: callout-tip
## Try it!

The `country` column still combines two pieces of information (`"Argentina (ARG)"`).

1.  Use `separate()` to split it into `country_name` and `country_code`
2.  By default, `separate()` splits at any non-letter, non-number character
3.  You may need the `extra = "drop"` argument to handle the trailing `)`

```{r}
#| eval: false
tidy_data <- separate(
  data = split_data,
  col = country,
  into = c("country_name", "country_code"),
  extra = "drop"
)
```
:::

# Pipes in Action {background-color="#2c3e50"}

## The Same Workflow, Without Pipes

Here's everything we did, written as separate steps:

```{r}
untidy <- read_csv("data/untidy.csv", show_col_types = FALSE)

long <- pivot_longer(untidy,
                     cols = `1950`:`2025`,
                     names_to = "year",
                     values_to = "women_members")

split_data <- separate(long,
                       col = women_members,
                       into = c("women", "members"),
                       sep = " / ",
                       convert = TRUE)

tidy_data <- separate(split_data,
                      col = country,
                      into = c("country_name", "country_code"),
                      extra = "drop")
```

. . .

Four intermediate objects. Easy to lose track of which one is "current."

## The Pipe Version

```{r}
tidy_data <- read_csv("data/untidy.csv", show_col_types = FALSE) |>
  pivot_longer(
    cols = `1950`:`2025`,
    names_to = "year",
    values_to = "women_members"
  ) |>
  separate(
    col = women_members,
    into = c("women", "members"),
    sep = " / ",
    convert = TRUE
  ) |>
  separate(
    col = country,
    into = c("country_name", "country_code"),
    extra = "drop"
  )
```

. . .

One object, read top-to-bottom: read it, then pivot, then separate, then separate again.

## How to Read a Pipe

1.  Read in the data with `read_csv()` — **and then**
2.  Reshape longer with `pivot_longer()` — **and then**
3.  Split `women_members` into two columns — **and then**
4.  Split `country` into two columns

. . .

::: callout-note
You can debug a pipe by highlighting from the start to **just before** the pipe you want to stop at, and running it. The output is the data at that step.
:::

## When to Break the Chain

Don't feel obligated to fit *every* step into one pipeline.

. . .

Break it up when:

-   The chain gets so long it's hard to read
-   You want to inspect or save an intermediate result
-   You're debugging and need to isolate a step

. . .

Pipes are a tool for clarity, not a contest.

# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   **Tidy data**: one observation per row, one variable per column, one value per cell
-   `pivot_longer()` stacks wide columns into a single column (great for plotting and summarising)
-   `pivot_wider()` does the reverse (great for display)
-   `separate()` splits a column with multiple values per cell into several columns
-   Chain tidying steps with `|>` to keep your code readable

. . .

Tidying is a conceptual skill — practice with real datasets is the only way to get fluent.
