---
title: "Data Relations"
subtitle: "Data Skills for Psychology — Module 7"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

These slides are adapted from [Applied Data Skills Ch. 5: Data Relations](https://psyteachr.github.io/ads-v3/05-joins.html).

## Today's Goals

By the end of this class, you will be able to:

-   Match related data across multiple tables using **joins**
-   Choose between `left_join()`, `right_join()`, `inner_join()`, and `full_join()`
-   Stack tables together with `bind_rows()` and `bind_cols()`
-   Use **set operations** to compare two tables
-   Diagnose and fix common join problems (mismatched types)

## Setup

Make sure you have the tidyverse loaded:

```{r}
library(tidyverse)
```

::: callout-tip
Grab the [Data transformation cheat sheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html) — it has a great visual summary of all the joins on the back page.
:::

# Why Combine Data? {background-color="#2c3e50"}

## Real data lives in many places

Almost every real-world analysis starts by combining data from multiple sources:

. . .

-   Patient records in one file, lab results in another
-   Survey responses split across months or sites
-   Student demographics in one sheet, grades in another
-   Trial-level behavioral data plus participant-level questionnaires

. . .

The **`*_join()`** family lets you stitch tables together using a shared identifier (like a participant ID).

## Two Example Tables

We'll use two tiny demo tables. First, a `patients` table:

```{r}
patients <- tibble(
  id = 1:5,
  city = c("Port Ellen", "Dufftown", NA,
           "Aberlour", "Tobermory"),
  postcode = c("PA42 7DU", "AB55 4DH", NA,
               "AB38 7RY", "PA75 6NR")
)
```

. . .

And a `visits` table — counts of GP visits per patient:

```{r}
visits <- tibble(
  id = c(2, 3, 4, 4, 5, 5, 6, 6, 7),
  visits = c(2, 1, 4, 0, 9, 11, 5, 0, 3)
)
```

## What's in each table?

::::: columns
::: {.column width="50%"}
**`patients`** (5 rows)

| id  | city       | postcode |
|-----|------------|----------|
| 1   | Port Ellen | PA42 7DU |
| 2   | Dufftown   | AB55 4DH |
| 3   | NA         | NA       |
| 4   | Aberlour   | AB38 7RY |
| 5   | Tobermory  | PA75 6NR |
:::

::: {.column width="50%"}
**`visits`** (9 rows)

| id  | visits |
|-----|--------|
| 2   | 2      |
| 3   | 1      |
| 4   | 4      |
| 4   | 0      |
| 5   | 9      |
| 5   | 11     |
| 6   | 5      |
| 6   | 0      |
| 7   | 3      |
:::
:::::

. . .

Notice: some IDs appear in only one table, and some IDs appear **multiple times** in `visits`.

# Mutating Joins {background-color="#2c3e50"}

## What is a Mutating Join?

A **mutating join** adds new columns from one table onto another, matching rows by a key.

. . .

The general syntax is the same for all four:

```{r}
#| eval: false
left_join(x, y, by = "id")
right_join(x, y, by = "id")
inner_join(x, y, by = "id")
full_join(x, y, by = "id")
```

. . .

-   `x` = the first (left) table
-   `y` = the second (right) table
-   `by` = which column(s) to match on

## `left_join()`

Keeps **all rows from the left table** and adds matching info from the right.

```{r}
left_data <- left_join(patients, visits, by = "id")
left_data
```

. . .

Result: **all 5 patients are kept**. Patients with multiple visits get **duplicated rows**. Patient 1 (no visits) gets `NA` in the `visits` column.

## `right_join()`

Keeps **all rows from the right table** and adds matching info from the left.

```{r}
right_data <- right_join(patients, visits, by = "id")
right_data
```

. . .

Result: **all 9 visits are kept**. IDs in `visits` that don't appear in `patients` (ids 6, 7) get `NA` for city and postcode.

. . .

::: callout-note
`right_join(x, y)` returns the same rows as `left_join(y, x)` — just with the columns in a different order.
:::

## `inner_join()`

Keeps **only rows that appear in both tables**.

```{r}
inner_data <- inner_join(patients, visits, by = "id")
inner_data
```

. . .

Result: only IDs 2, 3, 4, 5 remain. Patient 1 (no visits) is dropped. Visits for IDs 6 and 7 (not in `patients`) are also dropped.

## `full_join()`

Keeps **everything from both tables**, filling in `NA` where there's no match.

```{r}
full_data <- full_join(patients, visits, by = "id")
full_data
```

. . .

Result: every ID from either table appears. Patient 1 has `NA` for visits. Visits 6 and 7 have `NA` for city and postcode.

## Picking the Right Join {.smaller}

| Join               | What you keep                            |
|--------------------|------------------------------------------|
| `left_join(x, y)`  | All rows from `x`; matched info from `y` |
| `right_join(x, y)` | All rows from `y`; matched info from `x` |
| `inner_join(x, y)` | Only rows in **both**                    |
| `full_join(x, y)`  | Rows from **either**                     |

. . .

A good first question to ask: **which table is the "ground truth" I want to preserve?**

## Activity: Which Join?

::: callout-tip
## Choose the right join for each scenario

1.  You have `registrations` (everyone signed up with a doctor) and `jabs` (everyone who got a flu shot). You want to analyze outcomes for people who are **both** registered and vaccinated.

2.  You have `boards` (all NHS regions and their populations) and `attendances` (clinic visits, only some boards). You want **every** NHS board, with attendance filled in where available.

3.  You have `fuel_support` and `disability_adaptations`. You want **everyone** in either list, combining both sets of variables.

Think about it — which join, and why?
:::

# Chaining Joins {background-color="#2c3e50"}

## Joining More Than Two Tables

Joins are **two-table verbs**. To combine three+ tables, just chain them.

```{r}
satisfaction <- tibble(
  id = 1:5,
  satisfaction = c(4, 3, 2, 3, 1)
)
```

. . .

```{r}
# Using the pipe — read |> as "and then"
pipe_join <- patients |>
  left_join(visits, by = "id") |>
  left_join(satisfaction, by = "id")
```

. . .

Each `left_join` step adds new columns onto whatever came before.

## Always Check Your Output!

::: callout-warning
Joins are easy to write but surprisingly easy to **get wrong**. Before moving on, ask:

-   Does the result have **more or fewer rows** than expected?
-   Does it have the **right number of columns**?
-   Are there **NAs** where you didn't expect them?

If you build up a multi-step pipe, **check the result at each step.**
:::

. . .

Useful inspection functions: `glimpse()`, `summary()`, `nrow()`.

# Binding Joins {background-color="#2c3e50"}

## Binding ≠ Joining

**Joins** match rows by a key column.

**Binding** just stacks tables — rows on top of rows, or columns side by side. There's no matching logic.

. . .

-   `bind_rows()` → stack rows (must have same column names)
-   `bind_cols()` → stack columns (must have same number of rows, in the same order)

## `bind_rows()`

Add more patients to the existing table:

```{r}
new_patients <- tibble(
  id = 6:9,
  city = c("Falkirk", "Ardbeg", "Doogal", "Kirkwall"),
  postcode = c("FK1 4RS", "PA42 7EA",
               "G81 4SJ", "KW15 1SE")
)

bindr_data <- bind_rows(patients, new_patients)
bindr_data
```

. . .

Columns don't have to be in the same **order** — they just need the same **names**. Missing columns get filled with `NA`.

## `bind_cols()`

When the tables are in the **same row order**, you can paste columns side by side:

```{r}
new_info <- tibble(
  smoker = c("no", "no", "yes", "yes", "yes")
)

bindc_data <- bind_cols(patients, new_info)
bindc_data
```

. . .

::: callout-warning
`bind_cols()` is dangerous! It assumes rows are aligned. If the order is off, your data will be silently misaligned. Whenever possible, use a **mutating join** with a shared ID instead.
:::

# Set Operations {background-color="#2c3e50"}

## Comparing Whole Rows

Set operations compare two tables and return rows that:

-   appear in **both** → `intersect()`
-   appear in **either** → `union()` (deduplicated)
-   appear in one but **not the other** → `setdiff()`

. . .

Unlike joins, set operations compare **entire rows**, not just key columns. The tables need the same column names.

## `intersect()` and `union()`

```{r}
new_patients <- tibble(
  id = 5:9,
  postcode = c("PA75 6NR", "FK1 4RS", "PA42 7EA",
               "G81 4SJ", "KW15 1SE"),
  city = c("Tobermory", "Falkirk", "Ardbeg",
           "Doogal", "Kirkwall")
)

# rows that appear in BOTH tables
intersect(patients, new_patients)

# rows in EITHER table, with duplicates removed
union(patients, new_patients)
```

. . .

::: callout-note
`union()` is like `bind_rows()` but it removes exact duplicate rows.
:::

## `setdiff()`

Rows in the first table that are **not** in the second:

```{r}
setdiff(patients, new_patients)
```

. . .

Order matters! `setdiff(A, B)` and `setdiff(B, A)` answer different questions.

. . .

::: callout-warning
There are also `base::intersect()`, `base::union()`, and `base::setdiff()` that behave differently. Always load the tidyverse so the `dplyr::` versions are used.
:::

## Activity: Bind or Set Operation?

::: callout-tip
## Pick the right tool

1.  Two yearly survey extracts (`survey_2023`, `survey_2024`) — you want one combined dataset, **keeping duplicates** if a respondent appears in both years.

2.  Two extracts of an equality register — you want a combined dataset **with duplicates removed**.

3.  A master staff list and an HR export — you want to find staff in the master list that are **missing** from the export.

4.  A `patients` table and a `predictions` table with the same number of rows in the same order — you want to glue the prediction column on.
:::

# Putting It Together {background-color="#2c3e50"}

## A Realistic Workflow

Imagine you have:

-   `grades1` — exam scores (one row per student, has an `id` and `Points`)
-   `grades2` — essay scores (different students)
-   `schedule` — lookup table mapping `Points` → letter `Grade`

. . .

You want:

1.  Every student joined to their letter grade
2.  A "completes" dataset of students with **both** an exam and an essay grade

## The Pipeline {.smaller}

```{r}
# add letter grades to each table
exam_all  <- left_join(grades1, schedule, by = "Points")
essay_all <- left_join(grades2, schedule, by = "Points")

# only students with a non-missing grade
exam_grades  <- inner_join(grades1, schedule, by = "Points")
essay_grades <- inner_join(grades2, schedule, by = "Points")

# students with BOTH an exam and an essay grade
completes <- inner_join(exam_grades, essay_grades,
                        by = "id",
                        suffix = c("_exam", "_essay")) |>
  select(id, Grade_exam, Grade_essay)
```

. . .

Notice how `suffix` lets you rename columns that share a name between the two tables.

# Wrapping Up {background-color="#2c3e50"}

## Key Takeaways

-   **Mutating joins** add columns: `left_join`, `right_join`, `inner_join`, `full_join`

-   **Binding** stacks tables: `bind_rows` (more rows), `bind_cols` (more columns — be careful!)

-   **Set operations** compare whole rows: `intersect`, `union`, `setdiff`

-   **Always check** your result: did the row count change as expected? Any unexpected `NA`s?
