---
title: "Quarter Review"
subtitle: "Data Skills for Psychology — Spring 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    hash-type: number
    chalkboard: true
    menu: true
    progress: true
    controls: true
    navigation-mode: vertical
    code-line-numbers: false 
    code-overflow: wrap
    highlight-style: github
    title-slide-attributes:
      data-background-color: "#2c3e50"
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

Ten weeks of data skills — let's see how far we've come.

## What We'll Cover Today

A tour through every module this quarter:

1.  Programming Foundations
2.  Data Summaries
3.  Data Visualization
4.  Data Wrangling
5.  Data Tidying
6.  Data Relations
7.  Statistical Analysis
8.  Machine Learning

## Setup

```{r}
library(tidyverse)
library(janitor)
library(afex)
library(emmeans)
library(caret)
```

We'll use the **Faul et al. (2025)** emotional memory dataset throughout.

[Download `vivid_df_avg.csv`](../files/vivid_df_avg.csv)

# Programming Foundations {background-color="#2c3e50"}

## The Building Blocks

Every R program is built from five things:

. . .

| Concept       | What it does          | Example                         |
|---------------|-----------------------|---------------------------------|
| **Variables** | Name a value          | `n_participants <- 307`         |
| **Vectors**   | Store multiple values | `c(19, 23, 45, 67)`             |
| **Loops**     | Repeat code           | `for (s in subjects) { ... }`   |
| **Branches**  | Make decisions        | `if (age > 35) { ... }`         |
| **Functions** | Package reusable code | `my_fun <- function(x) { ... }` |

## Variables and Vectors

```{r}
# Single values
n_participants <- 307
task <- "emotional memory"

# Vectors — multiple values of the same type
ages <- c(19, 23, 45, 67, 34)
conditions <- c("Negative", "Positive", "Neutral")

# Selecting from a vector
ages[1]          # first value
ages[ages > 30]  # values matching a condition
```

. . .

::: callout-warning
Mixing types in a vector causes **coercion** — everything becomes a character.
:::

## Data Types

```{r}
typeof(307)        # double (numeric)
typeof("Negative") # character
typeof(TRUE)       # logical

is.numeric(307)    # TRUE
is.numeric("307")  # FALSE — quotes make it text!
```

. . .

**Factors** are a special type for ordered categories:

```{r}
age_group <- factor("Younger Age",
                    levels = c("Younger Age", "Middle Age", "Older Age"))
```

## Branches: `if` / `else`

```{r}
age <- 28

if (age < 35) {
  print("Younger Age group")
} else if (age < 65) {
  print("Middle Age group")
} else {
  print("Older Age group")
}
```

. . .

The condition inside `if ()` must evaluate to `TRUE` or `FALSE`. Only the matching block runs.

## Loops: `for`

```{r}
conditions <- c("Negative", "Positive", "Neutral")

for (cond in conditions) {
  subset <- df |> filter(Emotion == cond)
  cat(cond, "— mean vividness:", mean(subset$meanVivid), "\n")
}
```

. . .

For each value in `conditions`, the code inside `{ }` runs with `cond` set to that value.

. . .

::: callout-note
In practice, tidyverse functions like `group_by()` + `summarise()` often replace explicit loops — but loops are essential when you need more custom control.
:::

## Functions

```{r}
# Using a built-in function
mean(c(5.2, 4.8, 6.1, 3.9))

# Writing your own
sem <- function(x) {
  sd(x) / sqrt(length(x))
}

sem(ages)
```

. . .

Functions take **inputs** (arguments) and return an **output**. They're how you avoid copy-pasting the same code repeatedly.

## Activity: Write a Function

::: callout-tip
## Try it!

1.  Write a function called `zscore` that standardizes a vector: `(x - mean(x)) / sd(x)`
2.  Test it on BDI (depression) scores: `c(12, 5, 22, 8, 17, 3, 30)`
3.  **Bonus**: Add an `na.rm` argument that passes through to `mean()` and `sd()`
:::

# Data Summaries {background-color="#2c3e50"}

## Importing Data

```{r}
df <- read_csv("vivid_df_avg.csv")
```

. . .

Always check what you loaded:

```{r}
glimpse(df)   # column names, types, first values
nrow(df)      # number of rows
names(df)     # column names
```

. . .

This dataset has one row per participant per emotion condition — `Subject_ID`, `Emotion`, `meanVivid`, `meanAccuracy`, `Age`, `Age_Group`, `Gender`, `BDI_Sum`.

## The Pipe `|>`

The pipe sends the output of one function as the input to the next. Read it as **"and then."**

```{r}
# Without pipes — hard to follow
summarise(group_by(filter(df, Emotion == "Negative"),
                   Subject_ID), m = mean(meanVivid))

# With pipes — reads like instructions
df |>
  filter(Emotion == "Negative") |>
  group_by(Subject_ID) |>
  summarise(m = mean(meanVivid))
```

## Counting and Summarizing

```{r}
# Count rows per group
df |> count(Emotion)
df |> count(Age_Group, Emotion)

# Compute summary statistics
df |>
  group_by(Emotion) |>
  summarise(
    n        = n(),
    m_vivid  = mean(meanVivid, na.rm = TRUE),
    sd_vivid = sd(meanVivid, na.rm = TRUE)
  ) |>
  ungroup()
```

. . .

::: callout-warning
Always `ungroup()` when you're done — forgetting causes mysterious bugs later.
:::

## Activity: Summarize the Memory Data

::: callout-tip
## Try it!

1.  How many participants are in each `Age_Group`? Use `count()`
2.  What is the mean vividness (`meanVivid`) for each `Emotion` condition?
3.  What is the mean accuracy (`meanAccuracy`) for each combination of `Emotion` and `Age_Group`?
4.  **Bonus**: Which participant (`Subject_ID`) had the highest mean vividness across all conditions?
:::

# Data Visualization {background-color="#2c3e50"}

## The Grammar of Graphics

`{ggplot2}` builds plots in **layers**:

```{r}
ggplot(df, aes(x = Emotion, y = meanVivid)) +  # canvas + mapping
  geom_boxplot() +                               # geometry
  labs(title = "Vividness by Emotion") +         # labels
  theme_bw()                                     # theme
```

. . .

Every plot needs: **data**, **aesthetic mapping** (`aes()`), and at least one **geom**.

## Choosing the Right Geom

| Data                     | Good geoms                           |
|--------------------------|--------------------------------------|
| One continuous variable  | `geom_histogram()`, `geom_density()` |
| One categorical variable | `geom_bar()`                         |
| Continuous × categorical | `geom_boxplot()`, `geom_violin()`    |
| Continuous × continuous  | `geom_point()`, `geom_smooth()`      |

## Building Up Layers

```{r}
# Start simple
ggplot(df, aes(x = Emotion, y = meanVivid)) +
  geom_violin()
```

. . .

```{r}
# Add jittered points on top
ggplot(df, aes(x = Emotion, y = meanVivid)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.3)
```

. . .

```{r}
# Add color, labels, and facets
ggplot(df, aes(x = Emotion, y = meanVivid, fill = Emotion)) +
  geom_violin() +
  facet_wrap(~ Age_Group) +
  labs(x = "Emotion", y = "Mean Vividness Rating") +
  theme_bw() +
  theme(legend.position = "none")
```

## Customizing Plots 

```{r}
# Scatterplot with trendline
ggplot(df, aes(x = BDI_Sum, y = meanVivid)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x) +
  labs(x = "Depression Score (BDI)", y = "Mean Vividness")

# Save your plot
ggsave("vividness_by_emotion.png", width = 8, height = 5)
```

## Activity: Visualize Vividness

::: callout-tip
## Try it!

1.  Plot the distribution of `meanVivid` as a histogram, colored by `Emotion`
2.  Add `facet_wrap(~ Emotion)` — does the shape differ across conditions?
3.  Make a scatterplot of `Age` (x) vs. `meanVivid` (y)
4.  **Bonus**: Add `geom_smooth(method = "lm")` — is there a relationship with age?
:::

# Data Wrangling {background-color="#2c3e50"}

## The Six Core  Verbs

| Verb          | What it does                       |
|---------------|------------------------------------|
| `select()`    | Keep or drop **columns**           |
| `filter()`    | Keep rows matching a **condition** |
| `arrange()`   | **Sort** rows                      |
| `mutate()`    | Create or transform **columns**    |
| `summarise()` | Collapse to **summary statistics** |
| `group_by()`  | Apply operations **within groups** |

## `select()` and `filter()`

```{r}
# Keep useful columns
df |> select(Subject_ID, Age, Age_Group, Emotion, meanVivid, meanAccuracy)

# Drop a column
df |> select(-meanErrorAbs)

# Filter rows
df |> filter(Age_Group == "Younger Age")
df |> filter(Emotion %in% c("Negative", "Neutral"))
df |> filter(!is.na(meanVivid))
```

## `mutate()` and `arrange()`

```{r}
# Create new columns
df |>
  mutate(
    vivid_z      = (meanVivid - mean(meanVivid)) / sd(meanVivid),
    accuracy_pct = meanAccuracy * 100
  )

# Sort
df |> arrange(desc(meanVivid))
df |> arrange(Subject_ID, Emotion)
```

## `case_when()` — Multiple Conditions

```{r}
df |>
  mutate(
    bdi_category = case_when(
      BDI_Sum < 14  ~ "Minimal",
      BDI_Sum < 20  ~ "Mild",
      BDI_Sum < 29  ~ "Moderate",
      BDI_Sum >= 29 ~ "Severe"
    )
  )
```

. . .

`case_when()` is `if/else` for whole columns — much cleaner than nesting `ifelse()` calls.

## Activity: Wrangle the Memory Data

::: callout-tip
## Try it!

1.  Keep only rows where `Age_Group == "Younger Age"` and `meanVivid` is not missing
2.  Create a new column `vivid_z` that z-scores `meanVivid`
3.  Compute mean z-scored vividness by `Emotion`
4.  **Bonus**: Use `arrange(desc(...))` — which emotion condition has the highest vividness?
:::

# Data Tidying {background-color="#2c3e50"}

## What is Tidy Data?

Data is **tidy** when:

1.  Each **variable** has its own column
2.  Each **observation** has its own row
3.  Each **value** has its own cell

. . .

`vivid_df_avg.csv` is already tidy — one row per participant per emotion condition.

## Wide vs. Long Format

::::: columns
::: {.column width="50%"}
**Wide** (one row per participant)

| Subject_ID | Negative | Neutral | Positive |
|------------|----------|---------|----------|
| P001       | 5.2      | 4.1     | 3.8      |
| P002       | 6.1      | 3.9     | 5.0      |
:::

::: {.column width="50%"}
**Long** (one row per observation)

| Subject_ID | Emotion  | meanVivid |
|------------|----------|-----------|
| P001       | Negative | 5.2       |
| P001       | Neutral  | 4.1       |
| P001       | Positive | 3.8       |
| P002       | Negative | 6.1       |
:::
:::::

. . .

`ggplot2` and `dplyr` generally want **long** format.

## `pivot_longer()` and `pivot_wider()`

```{r}
# Long → Wide (e.g., one column per emotion condition)
df_wide <- df |>
  select(Subject_ID, Emotion, meanVivid) |>
  pivot_wider(
    names_from  = Emotion,
    values_from = meanVivid
  )

# Wide → Long (back to original shape)
df_long <- df_wide |>
  pivot_longer(
    cols      = c(Negative, Neutral, Positive),
    names_to  = "Emotion",
    values_to = "meanVivid"
  )
```

## `separate()`

```{r}
# Split "Younger Age" into "Younger" and "Age"
df |>
  separate(
    col  = Age_Group,
    into = c("group", "label"),
    sep  = " "
  )
```

. . .

Useful when one column contains two pieces of information mashed together.

# Data Relations {background-color="#2c3e50"}

## Joining Tables

When data about the same participants lives in separate files, we **join** them.

```{r}
# All four joins — the key is which rows to keep
inner_join(df_trials, df_demo, by = "Subject_ID")  # only matched rows
left_join(df_trials, df_demo, by = "Subject_ID")   # all trials, matched demographics
right_join(df_trials, df_demo, by = "Subject_ID")  # all demographics, matched trials
full_join(df_trials, df_demo, by = "Subject_ID")   # everything from both
```

. . .

::: callout-note
Use `by = c("subj" = "id")` when the key column has different names in each table.
:::

## Which Join to Use?

| Situation                                            | Join         |
|------------------------------------------------------|--------------|
| Only participants with complete data from both files | `inner_join` |
| All trials; add demographics where available         | `left_join`  |
| Combine all records, fill gaps with NA               | `full_join`  |

. . .

`left_join` is the most common in practice.

## Stacking Data

```{r}
# Combine datasets from multiple sessions or sites
bind_rows(df_session1, df_session2)

# Combine multiple participants' files into one
all_files <- list.files("data/", pattern = ".csv", full.names = TRUE)
df_all <- map(all_files, read_csv) |> bind_rows()
```

# Statistical Analysis {background-color="#2c3e50"}

## Descriptive vs. Inferential Statistics

**Descriptive** statistics summarize what's in your sample: means, SDs, counts.

. . .

**Inferential** statistics ask: *could this pattern have occurred by chance?*

-   We compute a **test statistic** and a **p-value**
-   p \< .05 means we reject the null hypothesis (by convention)
-   **Effect sizes** tell us *how big* the difference is

## t-tests

```{r}
# Separate conditions for a paired t-test
neg <- df |> filter(Emotion == "Negative")
neu <- df |> filter(Emotion == "Neutral")

# Paired t-test (same participants, two conditions)
t.test(neg$meanAccuracy, neu$meanAccuracy, paired = TRUE)
```

. . .

Report: t(df), p-value, and Cohen's d as the effect size.

## ANOVA with `afex`

```{r}
library(afex)
library(emmeans)

# One-way repeated-measures ANOVA
results <- aov_ez(
  id     = "Subject_ID",
  dv     = "meanVivid",
  within = "Emotion",
  data   = df
)
print(results)

# Follow-up pairwise comparisons
emmeans(results, pairwise ~ Emotion)
```

. . .

`aov_ez` automatically applies Greenhouse-Geisser correction and reports **generalized eta-squared** (ges) as the effect size.

## Regression

```{r}
# Does depression score predict memory vividness?
model <- lm(meanVivid ~ BDI_Sum + Age + Gender, data = df)
summary(model)
```

. . .

Look at: **coefficient estimates** (direction and size), **p-values** (significance), **R²** (variance explained).

## Activity: Test It

::: callout-tip
## Try it!

1.  Run a paired t-test comparing `meanAccuracy` between `Negative` and `Neutral` conditions
2.  Interpret: is the difference statistically significant? What's the direction?
3.  Visualize the result with `geom_boxplot()` + `geom_jitter()`
4.  **Bonus**: Run a repeated-measures ANOVA across all three emotion conditions with `aov_ez()`
:::

# Machine Learning {background-color="#2c3e50"}

## From Statistics to Prediction

Statistics asks: *is there a relationship?*

Machine learning asks: *can I predict new outcomes?*

. . .

Both use **regression** — but ML focuses on **generalizability**: does the model work on data it has never seen?

## Fit a Regression Model

```{r}
# Build a model predicting vividness from depression and age
model <- lm(meanVivid ~ BDI_Sum + Age, data = df)
summary(model)

# Generate predictions
df$predicted <- predict(model, newdata = df)
```

. . .

::: callout-warning
Evaluating a model on the same data it was trained on is **overly optimistic**. We need held-out data for a fair test.
:::

## Cross-Validation with `caret`

```{r}
library(caret)

# Set up 5-fold cross-validation
train_control <- trainControl(method = "cv", number = 5)

# Fit model with cross-validation
cv_model <- train(
  meanVivid ~ BDI_Sum + Age,
  data      = df,
  method    = "lm",
  trControl = train_control
)

cv_model   # R², RMSE, MAE averaged across folds
```

## Model Evaluation

| Metric   | Meaning                                                     |
|----------|-------------------------------------------------------------|
| **R²**   | Proportion of variance explained (0–1, higher = better)     |
| **RMSE** | Average prediction error in original units (lower = better) |
| **MAE**  | Mean absolute error (lower = better)                        |

. . .

Cross-validation gives a **more honest** estimate than fitting and testing on the same data.

# Putting It All Together {background-color="#2c3e50"}

## The Full Data Science Pipeline

Every project this quarter followed the same arc:

. . .

```         
Import → Inspect → Tidy → Wrangle → Visualize → Analyze → Communicate
```

. . .

These aren't separate skills — they're one continuous workflow.

## The Tidyverse Ecosystem

| Package            | What you used it for                      |
|--------------------|-------------------------------------------|
| `readr`            | `read_csv()` — import data                |
| `dplyr`            | `filter`, `mutate`, `summarise`, joins    |
| `tidyr`            | `pivot_longer`, `pivot_wider`, `separate` |
| `ggplot2`          | All visualization                         |
| `afex` / `emmeans` | ANOVA and follow-up tests                 |
| `caret`            | Cross-validated model fitting             |

## A Complete Example

```{r}
# Wrangle
df_neg <- df |>
  filter(Emotion == "Negative") |>
  mutate(bdi_category = case_when(
    BDI_Sum < 14  ~ "Minimal",
    BDI_Sum < 20  ~ "Mild",
    BDI_Sum >= 20 ~ "Moderate/Severe"
  ))

# Visualize
ggplot(df_neg, aes(x = bdi_category, y = meanVivid)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.4) +
  labs(title = "Negative Memory Vividness by Depression Level",
       x = "BDI Category", y = "Mean Vividness") +
  theme_bw()

# Analyze
model <- lm(meanVivid ~ BDI_Sum + Age, data = df_neg)
summary(model)
```


# Wrapping Up {background-color="#2c3e50"}

## What You Can Now Do

-   **Import and inspect** real datasets in R
-   **Wrangle** messy data into analysis-ready shape
-   **Visualize** distributions, comparisons, and relationships
-   **Test hypotheses** with t-tests, ANOVA, and regression
-   **Build predictive models** and evaluate them with cross-validation
-   Write **reproducible** analyses in Quarto

. . .

That's a full research workflow. All in one quarter!

## Other Key Concepts

-   **`|>`** makes complex operations readable — read it as "and then"
-   **Tidy data** (one variable per column, one observation per row) makes everything else easier
-   **Visualization** is not just for communication — it's for understanding your data
-   **p-values** answer "could this be chance?"; effect sizes answer "does it matter?"
-   **Cross-validation** is how you know whether your model actually generalizes

## What's Next

You now have the foundations to:

-   Finish your **final projects** (this week!)
-   Apply these skills in research settings — lab work, thesis, RA positions
-   Pick up **Python** faster — the concepts transfer directly
-   Go deeper: Bayesian stats, neuroimaging pipelines, more complex models

. . .

