---
title: "Putting It All Together — In-Class Workshop (KEY)"
author: "Maureen Ritchey"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    code-overflow: wrap
    df-print: kable
---

## Overview

Today's workshop puts together everything you've learned this quarter: importing and
exploring data, wrangling it into shape, visualizing patterns, and building a predictive
model. We'll work with the **Big Five Inventory (BFI)**, a real dataset included in the
`psych` package. It contains personality item responses from 2,800 participants, along
with demographic information.

The Big Five personality traits are:

- Agreeableness
- Conscientiousness
- Extraversion
- Neuroticism
- Openness

---

## Setup

Install any packages you don't have yet, then load them.

```{r}
#| message: false
# install.packages(c("psych", "caret"))  # run if needed

library(tidyverse)
library(psych)
library(caret)
```

---

## Part 1: Get to Know the Data

Load the dataset and convert it to a tibble. The `bfi` object is a data frame built into
the `psych` package — it contains 25 personality items (5 per trait), plus age, gender,
and education level. Here is a website with more detailed information: https://www.personality-project.org/r/html/bfi.html

```{r}
data(bfi)
bfi_df <- as_tibble(bfi, rownames = "id")
```

Take a first look:

```{r}
glimpse(bfi_df)
```

**Understanding the columns:**

- `A1`–`A5`: Agreeableness items (each rated 1–6)
- `C1`–`C5`: Conscientiousness items
- `E1`–`E5`: Extraversion items
- `N1`–`N5`: Neuroticism items
- `O1`–`O5`: Openness items
- `gender`: 1 = Male, 2 = Female
- `education`: 1 (some high school) to 5 (graduate degree)
- `age`: age in years

How much missing data is there?
What functions do you know that would be useful for checking whether there is missing data?

```{r}
# Count NAs per column
colSums(is.na(bfi_df))

# Or use summary() to see min/max/NA counts for every column
summary(bfi_df)
```


### Activity: Explore the data

1. How many rows and columns are in `bfi_df`?
2. What is the average age of participants? What is the range?
3. How many participants are male vs. female?

```{r}
# 1. Dimensions
nrow(bfi_df)       
ncol(bfi_df)

# 2. Age summary
mean(bfi_df$age, na.rm = TRUE)
range(bfi_df$age, na.rm = TRUE)

# 3. Gender counts
count(bfi_df, gender)
# gender 1 = Male, 2 = Female
```

---

## Part 2: Wrangle

Right now the data has individual item scores. We need to compute a **mean score** for
each of the five traits. We'll use `rowMeans()` inside `mutate()` to average across the
five items for each trait.

Careful! Some of the items are reverse-coded, meaning that low ratings mean a higher score for that trait. This is coded in `bfi.keys`:

```{r}
print(bfi.keys)
```

Now that you know how to score the data, how would you create a new variable for each trait?

```{r}
# Items flagged with a minus sign in bfi.keys are reverse-coded.
# On a 1–6 scale, the reverse of a score x is: 7 - x
#
# Reverse-coded items:
#   Agreeableness:    A1
#   Conscientiousness: C4, C5
#   Extraversion:     E1, E2
#   Neuroticism:      (none)
#   Openness:         O2, O5

bfi_df <- bfi_df |>
  mutate(
    # Step 1: reverse-code the flagged items
    A1r = 7 - A1,
    C4r = 7 - C4,
    C5r = 7 - C5,
    E1r = 7 - E1,
    E2r = 7 - E2,
    O2r = 7 - O2,
    O5r = 7 - O5
  ) |>
  mutate(
    # Step 2: compute mean scale scores using corrected items
    Agreeableness     = rowMeans(cbind(A1r, A2, A3, A4, A5), na.rm = TRUE),
    Conscientiousness = rowMeans(cbind(C1, C2, C3, C4r, C5r), na.rm = TRUE),
    Extraversion      = rowMeans(cbind(E1r, E2r, E3, E4, E5), na.rm = TRUE),
    Neuroticism       = rowMeans(cbind(N1, N2, N3, N4, N5), na.rm = TRUE),
    Openness          = rowMeans(cbind(O1, O2r, O3, O4, O5r), na.rm = TRUE)
  )
```

Once you have all five scores, recode `gender` from 1/2 to labeled categories. Hint: use the factor command.

```{r}
bfi_df <- bfi_df |>
  mutate(gender = factor(gender, levels = c(1, 2), labels = c("Male", "Female")))
```

Finally, remove rows with missing values on any of the five scale scores. We need
complete cases for modeling later.

```{r}
bfi_complete <- bfi_df |>
  filter(
    !is.na(Agreeableness),
    !is.na(Conscientiousness),
    !is.na(Extraversion),
    !is.na(Neuroticism),
    !is.na(Openness)
  )

nrow(bfi_complete)   # should be ~2436
```

---

## Part 3: Visualize

### Distribution of Neuroticism

We'll start with Neuroticism since that's what we'll predict later. Use `ggplot` to visualize the distribution across the sample.

```{r}
ggplot(bfi_complete, aes(x = Neuroticism)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Neuroticism scores",
       x = "Mean Neuroticism (1–6)",
       y = "Count") +
  theme_minimal()
```

### Neuroticism by gender

Now split by gender.

```{r}
ggplot(bfi_complete, aes(x = gender, y = Neuroticism, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Neuroticism scores by gender",
       x = NULL, y = "Mean Neuroticism (1–6)") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Activity: Visualize a relationship

Choose one of the other Big Five traits and explore its relationship with Neuroticism.
Make a scatterplot with a smoothed trend line using `geom_smooth()`.

1. Do higher scores on your chosen trait tend to go along with higher or lower Neuroticism?
2. Does the relationship look linear?
3. **Bonus:** Color the points by gender. Does the pattern differ?

```{r}
# Example using Agreeableness — expect a negative relationship with Neuroticism
ggplot(bfi_complete, aes(x = Agreeableness, y = Neuroticism)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "steelblue") +
  labs(title = "Agreeableness vs. Neuroticism",
       x = "Mean Agreeableness (1–6)",
       y = "Mean Neuroticism (1–6)") +
  theme_minimal()

# Bonus: color by gender
ggplot(bfi_complete, aes(x = Agreeableness, y = Neuroticism, color = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = "Agreeableness vs. Neuroticism by gender") +
  theme_minimal()
```

---

## Part 4: Build a Predictive Model

Now let's use what you learned in the machine learning module. We'll predict **Neuroticism**
from the other four traits.

### Split into training and testing sets

```{r}
set.seed(42)

train_index <- createDataPartition(bfi_complete$Neuroticism, p = 0.8, list = FALSE)

train_data <- bfi_complete[train_index, ]
test_data  <- bfi_complete[-train_index, ]

nrow(train_data)
nrow(test_data)
```

### Fit a linear regression model with cross-validation

```{r}
ctrl <- trainControl(method = "cv", number = 5)

model <- train(
  Neuroticism ~ Agreeableness + Conscientiousness + Extraversion + Openness,
  data      = train_data,
  method    = "lm",
  trControl = ctrl
)

model
```

What does the cross-validated R² tell you? How much of the variation in Neuroticism can
the other four traits explain?

### Evaluate on the test set

```{r}
predictions <- predict(model, newdata = test_data)

postResample(pred = predictions, obs = test_data$Neuroticism)
```

### Look at the coefficients

```{r}
summary(model$finalModel)
```

Which trait is the strongest predictor of Neuroticism? Does the direction of the effect
make sense to you?

### Activity: Improve the model

Try at least one of the following:

1. **Add demographic predictors.** Include `age` and/or `education` alongside the four
   personality traits. Does R² go up?
2. **Try a different outcome.** Instead of predicting Neuroticism, predict Extraversion
   or Conscientiousness. Which one is easier to predict from the other traits?
3. **Bonus:** Try `method = "ridge"` instead of `"lm"` in the `train()` call (you may
   need `install.packages("glmnet")`). Does regularization help?

```{r}

```
